---
title: "USVI 2021 CINAR Temporal Metabolite Analysis"
author: "Brianna Garcia"
date: "`r Sys.Date()`"
output: html_document
---

# Document Set-up
## Set working directory
```{r setup, include=FALSE}

#Work PC directories
knitr::opts_knit$set(root.dir = "Z:/Brianna/Projects/2022_0328_CINAR_BC")
setwd("Z:/Brianna/Projects/2022_0328_CINAR_BC")

#Laptop directories
#knitr::opts_knit$set(root.dir = "/Volumes/KujLab/Brianna/Projects/2022_0328_CINAR_BC")
#setwd("/Volumes/KujLab/Brianna/Projects/2022_0328_CINAR_BC")

```

## Load required packages
```{r packages, message=FALSE}

library(RColorBrewer)
library(gridExtra)
library(sf)
library(ggspatial)
library(scales)
library(data.table)
library(ggdendro)
library(viridis)
library(vegan)
library(ggpubr)
library(tidyverse)
library(lubridate)
library(knitr)
library(kableExtra)
library(purrr)
library(patchwork)

```

## Define color palette for all downstream plots
```{r color palette}

CINAR_temporal_palette=c("#A1CAF1","#AE2337","#BFB386")

if(!exists("site_lookup", envir = .GlobalEnv)){
  site_lookup <- data.frame(site = c("LB_seagrass", "Tektite", "Yawzi", "control_extraction", "control_pcr", "control_seq"),
                            label = c("Lameshur Bay seagrass", "Tektite Reef", "Yawzi Reef",
                                      "Control (DNA Extraction)", "Control (PCR)", "Control (Sequencing)")) %>%
    dplyr::mutate(label = stringr::str_wrap(label, width = 16)) %>%
    tibble::deframe(.)
  site_colors <- pals::kelly(22)[6:(5+length(site_lookup))] %>%
    # site_colors <- viridisLite::cividis(n = length(site_lookup), direction = 1) %>%
    setNames(., names(site_lookup))
  sampling_time_lookup <- data.frame(sampling_time = c("dawn", "peak_photo"),
                                     label = c("Dawn", "Afternoon")) %>%
    tibble::deframe(.)
  sampling_time_colors <- pals::ocean.haline(n = length(sampling_time_lookup)) %>%
    setNames(., names(sampling_time_lookup))
  sampling_day_lookup <- data.frame(sampling_day = c("Day1", "Day2", "Day3", "Day4", "Day5"),
                                    label = c("20210122", "20210123", "20210124", "20210125", "20210126")) %>%
    tibble::deframe(.)
  sampling_day_colors <- pals::ocean.thermal(n = length(sampling_day_lookup)) %>%
    setNames(., names(sampling_day_lookup))
}


```

# Load in data and prep for statistical analysis

## Load data
```{r load data, message=FALSE}

data <- read_csv('Z:/Brianna/Projects/2022_0328_CINAR_BC/Analysis/MATLAB/BC_qaqc/20250204/temporal_mergedTable_metCols_CINARBC_20240204_BMG_manualCheck.csv')
data <- as_tibble(data,rownames="rowID_full")
data$rep_group<-as.factor(data$rep_group)
limits <- read_csv("Z:/Brianna/Projects/2022_0328_CINAR_BC/Analysis/MATLAB/BC_qaqc/20250204/CINAR_metsFiltered_LODLOQ.2025.02.04_manualCheck.csv")
seq <- read_csv("Z:/Brianna/Projects/2022_0328_CINAR_BC/Raw_data/sequence/mtab_CINAR_BC_032822_MATLABtrunc.csv")

```

## Merge in filenames
```{r merge filenames}

seq_pos <- seq%>%
  filter(ionMode == "pos", goodData == 1, sType == "rep" )%>%
  select(File_Name,sMatlabName)%>%
  rename(File_Name_Pos = File_Name)

seq_neg <- seq%>%
  filter(ionMode == "neg", goodData == 1, sType == "rep" )%>%
  select(File_Name,sMatlabName)%>%
  rename(File_Name_Neg = File_Name)

seq_mode_sidebyside <- merge(seq_neg,seq_pos,by = "sMatlabName")%>%
  rename(adaptedDervLabel = sMatlabName)

data_wFileName <- merge(seq_mode_sidebyside,data,by = "adaptedDervLabel",all.y = TRUE)

rm(seq_pos,seq_neg, seq_mode_sidebyside,data)

```

## Subset temporal data for larger dataset
```{r subset temporal data}

temporal <- data_wFileName %>%
  ungroup()%>%
  add_count(rep_group)%>%
  relocate(n,.after=rep_group)%>%
  filter(n > 1)%>%
  mutate(group = case_when(Site %in% c("Tektite","Yawzi") ~ "coral", 
                           Site == "LB_seagrass" ~ "seagrass"))%>%
  relocate(group, .after = Site)%>%
  rename(NPOC = NPOC_uM_, TN = TN_uM_)%>%
  mutate(Site = factor(Site, levels = c("LB_seagrass","Tektite","Yawzi")))%>%
  mutate(Site_abv =
           case_when(Site =="Yawzi" ~ "YZ",
                     Site =="Tektite" ~ "TK",
                     Site =="LB_seagrass" ~ "SG"),.after = Site)%>%
  mutate(Site_abv = factor(Site_abv, levels = c("YZ", "TK","SG")))%>%
  mutate(Date = parse_date_time(Date, orders = "mdy"),
         date_time = ymd_hms(paste(Date, Time)))

temporal_long <- temporal %>%
  pivot_longer(cols="3'AMP":"valine",names_to="metabolites",values_to="concentration",values_drop_na=TRUE)

mtabName_wMode <- limits$Molecule

limits <- limits%>%
  rename(metabolites = SimpleName)%>%
  select(-Molecule)

temporal_long_wLimits <- merge(temporal_long,limits,by="metabolites",all.y = TRUE)

rm(temporal_long)

temporal.mat <- temporal
rownames(temporal.mat) <- temporal$adaptedDervLabel
temporal.mat <- temporal.mat[,47:91]
temporal.mat <- temporal.mat[, ! apply(temporal.mat , 2 , function(x) 
  sd(x, na.rm = TRUE)==0 ) ] #remove metabolites with a SD of zero 

x <- colnames(temporal.mat) #pull out metabolite names

temporal_long_wLimits <- temporal_long_wLimits%>%
  ungroup()%>%
  filter(metabolites %in% x,
         !is.na(LOD_nM))%>%
  mutate(QuantFlag = concentration < LOQ_nM,
         LODflag = concentration < LOD_nM)%>%
  rename(LOD = LOD_nM, LOQ = LOQ_nM)%>%
  mutate(site_time = paste0(Site,"_",sample_type))

x <- unique(temporal_long_wLimits$metabolites)

temporal_wide <- temporal_long_wLimits %>%
  select(-c(LOD,LOQ,LODflag,QuantFlag))%>%
  pivot_wider(names_from = metabolites,values_from = concentration)

temporal.mat <- as.matrix(temporal_wide[,49:93])
rownames(temporal.mat) <- temporal_wide$adaptedDervLabel

rm(limits,data_wFileName,temporal)

```
# Outlier detection
## Calculate Euclidean distances by replicate group
```{r outlier distances}

# Create a plotting area for boxplots
par(mfrow = c(2, 3))  # Adjust rows/columns to fit all groups
par(mar = c(10, 4, 4, 2))  # Bottom, Left, Top, Right

# Iterate through each site_time group
unique_site_times <- unique(temporal_long_wLimits$site_time)

for (site in unique_site_times) {
  # Filter data for the current site_time
  site_data <- temporal_long_wLimits %>%
    filter(site_time == site) %>%
    select(metabolites,concentration,adaptedDervLabel,site_time)%>%
    pivot_wider(names_from = metabolites, values_from = concentration)%>%
    select(-site_time)

  # Convert to matrix for distance calculation
  site_matrix <- as.matrix(site_data[,-1])
  rownames(site_matrix) <- site_data$adaptedDervLabel
  
  # Compute pairwise Euclidean distances
  pairwise_distances <- dist(site_matrix, method = "euclidean")
  distance_matrix <- as.matrix(pairwise_distances)  # Convert to square form
  
  # Calculate mean distances for each sample
  mean_distances <- rowMeans(distance_matrix)
  
  # Determine thresholds for outliers using IQR
  iqr_distances <- IQR(mean_distances, na.rm = TRUE)
  lower_threshold <- quantile(mean_distances, 0.25, na.rm = TRUE) - 1.5 * iqr_distances
  upper_threshold <- quantile(mean_distances, 0.75, na.rm = TRUE) + 1.5 * iqr_distances
  
  # Plot boxplot for the distance matrix
  boxplot(
    distance_matrix, 
    main = paste("Site-Time:", site), 
    ylab = "Euclidean Distance", 
    col = "lightgray", 
    outline = FALSE, 
    las = 2
  )
  
  # Add points for mean distances
  points(mean_distances, col = "blue", pch = 21, size= 3, bg = "green")
  
  # Add lines for thresholds
  abline(h = c(lower_threshold, upper_threshold), col = "red", lty = 2, lwd = 2)
}

rm(unique_site_times,distance_matrix,site_data,site_matrix,iqr_distances,
   lower_threshold,upper_threshold,mean_distances,pairwise_distances,
   site)

```

## Outlier detection using NMDS with 95% CI
```{r outlier NMDS}

# Output plot list
plots <- list()

# Loop through each site
for (site in unique(temporal_wide$Site)) {
  # Filter data for the current site
  site_meta <- temporal_wide %>% filter(Site == site)
  site_data <- temporal.mat[rownames(temporal.mat) %in% site_meta$adaptedDervLabel, ]
  
  # Run NMDS
  nmds <- metaMDS(site_data, k = 2, trymax = 100, distance = "bray")
  
  # Extract scores
  scores <- as.data.frame(scores(nmds, display = "sites"))
  scores <- cbind(scores, site_meta)
 
  # Create the plot
  library(ggrepel)
  
  p <- ggscatter(scores, x = "NMDS1", y = "NMDS2", color = "sample_type", size = 3) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_vline(xintercept = 0, linetype = 2) +
    ggtitle(paste("NMDS for Site:", site)) +
    theme_bw(base_size = 12) +
    # Add 95% confidence ellipses
    stat_ellipse(aes(color = sample_type, fill = sample_type), level = 0.95, geom = "polygon", alpha = 0.05, show.legend = FALSE) +
    # Add labels
    geom_text(aes(label = adaptedDervLabel), vjust = -0.5, size = 3)
  
  
  # Save the plot
  plots[[site]] <- p
}

# View plots
#plots[[3]]  # Example for the first site

# Save all plots
#pdf(file= "Z:/Brianna/Projects/2022_0328_CINAR_BC/BC_CoralSeagrass_Temporal/figures/CINAR_USVI2021_NMDS_by_site_wEllipses.pdf",width = 11, height = 8.5,onefile=TRUE)
for (plot in plots) print(plot)
#dev.off()

```
## Remove outliers from dataset 
```{r remove outliers}

temporal_long_wLimits_outliersRmv <- temporal_long_wLimits%>%
  ungroup()%>%
  filter(adaptedDervLabel!= "CINAR_BC_73")%>%
  mutate(log2_concen = log2(concentration+1))

temporal_wide_outliersRmv_log2 <- temporal_long_wLimits_outliersRmv%>%
  select(-c(concentration,QuantFlag,LODflag,LOD,LOQ))%>%
  pivot_wider(names_from = "metabolites",values_from = "log2_concen")

temporal.mat_outliersRmv_log2 <- temporal_wide_outliersRmv_log2%>%
  select(c("3'AMP":"valine"))%>%
  as.matrix()

rownames(temporal.mat_outliersRmv_log2) <- temporal_wide_outliersRmv_log2$adaptedDervLabel

```

## Save table with outliers removed
```{r save outlier rmvd table}

#write.csv(temporal_long_wLimits_outliersRmv,file = "Z:/Brianna/Projects/2022_0328_CINAR_BC/BC_CoralSeagrass_Temporal/USVI_2021_CINAR_temporal_R_Script/csv outputs/USVI2021_CINARtemporal_BzCl_Exometabolite_QCd_longFormat_outliersRmvd_wMetadata_20250204.csv")

#write.csv(temporal.mat_outliersRmv_log2,file = "Z:/Brianna/Projects/2022_0328_CINAR_BC/BC_CoralSeagrass_Temporal/USVI_2021_CINAR_temporal_R_Script/csv outputs/USVI2021_CINARtemporal_BzCl_Exometabolite_QCd_wideFormat_outliersRmvd_log2_noMetadata_20250204.csv")

```
#Data summary

## HOBO Logger (Light and Temperature)
Note: Why does Yawzi reef have such lower lumens in the afternoon? Sampling order appears to be TK, YZ, and SG.
```{r HOBO}

temporal_long_wLimits_outliersRmv%>%
  group_by(Site,sample_type,Sampling_Day)%>%
  ggplot(aes(x=Sampling_Day,y=HOBO_lumens,fill=Site,shape = sample_type))+
  geom_point(size=3)+
  scale_shape_manual(values = c(21,22))+
  scale_fill_manual(values = site_colors, labels = site_lookup, breaks = names(site_lookup))+
  theme_bw()+
  facet_wrap(~Site)

temporal_long_wLimits_outliersRmv%>%
  group_by(Site,sample_type)%>%
  ggplot(aes(x=Site,y=HOBO_lumens,fill=Site,shape = sample_type))+
  geom_point(size = 3)+
  facet_wrap(~sample_type,scales = "free")+
  scale_fill_manual(values = site_colors, labels = site_lookup, breaks = names(site_lookup))+
  scale_shape_manual(values = c(21,22))+
  theme_bw()

temporal_long_wLimits_outliersRmv %>%
  group_by(sample_type, Site) %>%
  summarise(
    across(
      c(HOBO_lumens, HOBO_temp_F), 
      list(mean = ~mean(.x, na.rm = TRUE), sd = ~sd(.x, na.rm = TRUE)), 
      .names = "{.col}_{.fn}"
    )
  ) %>%
  mutate(
    HOBO_lumens = paste0(round(HOBO_lumens_mean, 2), " ± ", round(HOBO_lumens_sd, 2)),
    HOBO_temp_F = paste0(round(HOBO_temp_F_mean, 2), " ± ", round(HOBO_temp_F_sd, 2))
  ) %>%
  select(sample_type, Site, HOBO_lumens, HOBO_temp_F)

```

## NPOC and TN

### Average concentrations
```{r NPOC TN}

waterChem_long <- temporal_long_wLimits_outliersRmv%>%
  select(adaptedDervLabel,sample_type,Site,NPOC,TN,Sampling_Day, date_time)%>%
  pivot_longer(cols = c(NPOC, TN), names_to = "water_chem", values_to = "concen_uM")


waterChem_long %>%
  group_by(sample_type, Site, water_chem) %>%
  summarise(mean = mean(concen_uM, na.rm = TRUE), 
            sdev = sd(concen_uM, na.rm = TRUE))%>%
  mutate(mean_dev = paste0(round(mean, 2), " ± ", round(sdev, 2)))%>%
  select(sample_type, Site, mean_dev,water_chem)%>%
  pivot_wider(names_from = water_chem, values_from = mean_dev)

waterChem_long %>%
  group_by(sample_type,water_chem) %>%
  summarise(mean = mean(concen_uM, na.rm = TRUE), 
            sdev = sd(concen_uM, na.rm = TRUE))%>%
  mutate(mean_dev = paste0(round(mean, 2), " ± ", round(sdev, 2)))%>%
  select(sample_type,mean_dev,water_chem)%>%
  pivot_wider(names_from = water_chem, values_from = mean_dev)


```
### Boxplots by Site/Time

```{r Water chem boxplots}

waterChem_long%>%
  group_by(Site,sample_type)%>%
  ggplot(aes(x=Site,y=concen_uM,color=Site,linetype = sample_type))+
  geom_boxplot()+
  facet_wrap(~water_chem, scales="free", ncol=2)+
  theme_bw()+
  scale_color_manual(values = site_colors, labels = site_lookup, breaks = names(site_lookup))

```

### Plot DOC and TN over time 
``` {r doc tn over time}

watechem_overTime <- waterChem_long%>%
 mutate(sample_type = case_when(sample_type == "peak_photo" ~ "afternoon", 
                                 TRUE ~ sample_type),
          day_time = paste(Sampling_Day,sample_type),
          day_time = factor(day_time,levels = c("Day2 dawn", "Day2 afternoon", 
                                                "Day3 dawn", "Day3 afternoon",
                                                "Day4 dawn", "Day4 afternoon",
                                                "Day5 dawn", "Day5 afternoon")))%>%
  group_by(Site, day_time, water_chem) %>%
  summarize(
    mean_concentration = mean(concen_uM, na.rm = TRUE),
    sd_concentration = sd(concen_uM, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = day_time, y = mean_concentration, fill = Site)) + 
  geom_errorbar(aes(ymin = mean_concentration - sd_concentration,
                    ymax = mean_concentration + sd_concentration), 
                width = 0.2) +
  geom_point(size = 3, shape = 21, color = "black") +
  facet_grid(water_chem ~ Site, scales = "free")+
  scale_fill_manual(values = site_colors, labels = site_lookup, breaks = names(site_lookup))+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

watechem_overTime

```

### KW across sampling days (within site)
```{r KW across days}

# Function to perform Kruskal-Wallis test and FDR adjustment for a given subset of data
KW_days_waterchem <- function(data, site_filter, sample_type_filter) {
  data %>%
    ungroup() %>%
    filter(Site == site_filter, sample_type == sample_type_filter) %>%
    group_by(water_chem) %>%
    summarise(
      Kruskal_p = kruskal.test(concen_uM ~ Sampling_Day)$p.value,
      .groups = "drop"
    ) %>%
    mutate(
      FDR_adjusted_p = p.adjust(Kruskal_p, method = "fdr"),
      sig_05_fdr = FDR_adjusted_p < 0.05
    )
}

# List of subsets with corresponding labels
subset_list <- list(
  kw_SG_peak = list(site = "LB_seagrass", sample_type = "peak_photo"),
  kw_SG_dawn = list(site = "LB_seagrass", sample_type = "dawn"),
  kw_TK_peak = list(site = "Tektite", sample_type = "peak_photo"),
  kw_TK_dawn = list(site = "Tektite", sample_type = "dawn"),
  kw_YZ_peak = list(site = "Yawzi", sample_type = "peak_photo"),
  kw_YZ_dawn = list(site = "Yawzi", sample_type = "dawn")
)

# Combine all results into one summary table
set.seed(123)
summary_table_1 <- purrr::imap_dfr(
  subset_list,
  ~ KW_days_waterchem(waterChem_long, .x$site, .x$sample_type) %>%
      mutate(site_sample = .y)
)

# View the summary table
print(summary_table_1)

```
### Pairwise Wilcoxon Rank Sum Test
```{r water chem Site wilcoxon rank sum test}

# Perform pairwise Wilcoxon tests for each significant metabolite

set.seed(123)
pairwise_results <- waterChem_long %>%
  group_by(water_chem,sample_type) %>%
  summarise(
    pairwise_wilcox = list(
      pairwise.wilcox.test(
        x = concen_uM,
        g = Site,
        p.adjust.method = "fdr"
      )
    ),
    .groups = "drop"
  )

# View results
pairwise_results <- pairwise_results %>%
  mutate(pairwise_summary = map(pairwise_wilcox, ~ broom::tidy(.x)))

# Optionally, flatten for easier inspection
pairwise_summary <- pairwise_results %>%
  unnest(pairwise_summary)%>%
  mutate(significant = p.value < 0.05)%>%
  group_by(water_chem)

wilcoxon_waterChem_dawn_summary <- pairwise_summary

# Print results
print(wilcoxon_waterChem_dawn_summary)

```
### Kruskal Wallis within Site
```{r water chem KW sampling time}

# Function to perform Kruskal-Wallis test and FDR adjustment for a given site
KW_waterchem_fdr <- function(data, site_filter) {
waterChem_long %>%
    ungroup() %>%
    filter(Site == site_filter) %>%
    group_by(water_chem) %>%
    summarise(
      Kruskal_p = kruskal.test(concen_uM ~ sample_type)$p.value,
      .groups = "drop"
    ) %>%
    mutate(
      FDR_adjusted_p = p.adjust(Kruskal_p, method = "fdr"),
      sig_05_fdr = FDR_adjusted_p < 0.05
    )
}

# List of subsets with corresponding labels
subset_list <- list(
  kw_SG = list(site = "LB_seagrass"),
  kw_TK = list(site = "Tektite"),
  kw_YZ = list(site = "Yawzi")
)

# Combine all results into one summary table
set.seed(123)
summary_table_2 <- purrr::imap_dfr(
  subset_list,
  ~ KW_waterchem_fdr(waterChem_long, .x$site) %>%
      mutate(site = .x$site) # Use .x$site to extract the site name
)

# View the summary table
print(summary_table_2)

```


## Flow cytometry
```{r flow cytrometry}

temporal_long_wLimits_outliersRmv%>%
  pivot_longer(cols = c(Prochlorococcus, Synechococcus,Unpigmented_cells,Picoeukaryotes), names_to = "fcm_data", values_to = "counts")%>%
  group_by(Site,sample_type)%>%
  select(Site,sample_type,fcm_data,counts)%>%
  drop_na()%>%
  ggplot(aes(x=Site,y=counts,color=Site,linetype = sample_type))+
  geom_boxplot()+
  facet_wrap(~fcm_data, scales="free",ncol=1)+
  theme_bw()+
  scale_color_manual(values = site_colors, labels = site_lookup, breaks = names(site_lookup))

temporal_long_wLimits_outliersRmv %>%
  group_by(sample_type, Site) %>%
  summarise(
    across(
      c(Prochlorococcus, Synechococcus, Unpigmented_cells, Picoeukaryotes), 
      list(mean = ~mean(.x, na.rm = TRUE), sd = ~sd(.x, na.rm = TRUE)), 
      .names = "{.col}_{.fn}"
    )
  ) %>%
  mutate(
    Prochlorococcus = paste0(round(Prochlorococcus_mean, 2), " ± ", round(Prochlorococcus_sd, 2)),
    Synechococcus = paste0(round(Synechococcus_mean, 2), " ± ", round(Synechococcus_sd, 2)),
    Unpigmented_cells = paste0(round(Unpigmented_cells_mean, 2), " ± ", round(Unpigmented_cells_sd, 2)),
    Picoeukaryotes = paste0(round(Picoeukaryotes_mean, 2), " ± ", round(Picoeukaryotes_sd, 2))
  ) %>%
  select(sample_type, Site, Prochlorococcus, Synechococcus, Unpigmented_cells, Picoeukaryotes)

```

### FCM over time 
```{r fcm over time}


fcm_overTime <- temporal_long_wLimits_outliersRmv%>%
  pivot_longer(cols = c(Prochlorococcus, Synechococcus,Unpigmented_cells,Picoeukaryotes), names_to = "fcm_data", values_to = "counts")%>%
 mutate(sample_type = case_when(sample_type == "peak_photo" ~ "afternoon", 
                                 TRUE ~ sample_type),
          day_time = paste(Sampling_Day,sample_type),
          day_time = factor(day_time,levels = c("Day2 dawn", "Day2 afternoon", 
                                                "Day3 dawn", "Day3 afternoon",
                                                "Day4 dawn", "Day4 afternoon",
                                                "Day5 dawn", "Day5 afternoon")))%>%
  group_by(Site, day_time, fcm_data) %>%
  summarize(
    mean_counts = mean(counts, na.rm = TRUE),
    sd_counts = sd(counts, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = day_time, y = mean_counts, fill = Site)) + 
  geom_errorbar(aes(ymin = mean_counts - sd_counts,
                    ymax = mean_counts + sd_counts), 
                width = 0.2) +
  geom_point(size = 3, shape = 21, color = "black") +
  facet_grid(fcm_data ~ Site, scales = "free")+
  scale_fill_manual(values = site_colors, labels = site_lookup, breaks = names(site_lookup))+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

fcm_overTime

```

## Metabolite counts and concentration ranges
```{r mtab summary}

temporal_long_wLimits_outliersRmv %>%
  ungroup()%>%
  mutate(concen_forTable = case_when(concentration < LOD ~ NA, TRUE ~ concentration))%>%
  select(metabolites, Site, sample_type,concen_forTable, LOD)%>%
  mutate(sample_type = case_when(sample_type == "peak_photo" ~ "afternoon", 
                                 TRUE ~ sample_type))%>%
  group_by(metabolites, Site) %>%
  summarise(
    LOD = mean(LOD, na.rm = TRUE), # Use mean or another method to summarize LOD
    median_concentration = median(concen_forTable, na.rm = TRUE))%>%
  group_by(Site)%>%
  summarise(aboveLOD = sum(!is.na(median_concentration)))

temporal_long_wLimits_outliersRmv %>%
  ungroup()%>%
  mutate(concen_forTable = case_when(concentration < LOD ~ NA, TRUE ~ concentration))%>%
  select(metabolites, Site, sample_type,concen_forTable, LOD)%>%
  mutate(sample_type = case_when(sample_type == "peak_photo" ~ "afternoon", 
                                 TRUE ~ sample_type))%>%
  group_by(metabolites, Site, sample_type) %>%
  summarise(
    LOD = mean(LOD, na.rm = TRUE), # Use mean or another method to summarize LOD
    median_concentration = median(concen_forTable, na.rm = TRUE))%>%
  group_by(Site, sample_type)%>%
  summarise(aboveLOD = sum(!is.na(median_concentration)))

temporal_long_wLimits_outliersRmv %>%
  ungroup()%>%
  mutate(concen_forTable = case_when(concentration < LOD ~ NA, TRUE ~ concentration))%>%
  select(metabolites, Site, sample_type,concen_forTable, LOD)%>%
  summarise(min_conc = min(concen_forTable,na.rm = TRUE),
            max_conc = max(concen_forTable, na.rm = TRUE))

temporal_long_wLimits_outliersRmv %>%
  ungroup()%>%
  mutate(concen_forTable = case_when(concentration < LOD ~ NA, TRUE ~ concentration))%>%
  select(metabolites, Site, sample_type,concen_forTable, LOD)%>%
  arrange(concen_forTable)%>%
  view()

```

#Statistical tests
## Bray-Curtis Dissimilarity Calculations
```{r bray curtis dissimilitaries}

dist_usvi_mtab.d <- vegan::vegdist(temporal.mat_outliersRmv_log2, method = "bray", binary = FALSE, na.rm = TRUE)

dist_usvi_mtab.df <- vegan::betadisper(dist_usvi_mtab.d, temporal_wide_outliersRmv_log2$Site) %>%
  purrr::pluck("distances") %>%
  tibble::enframe(value = "dissimilarity", name = "adaptedDervLabel") %>%
  dplyr::left_join(., (temporal_wide_outliersRmv_log2 %>%
                         dplyr::select(adaptedDervLabel, sample_type, Date, Time, Sampling_Day, Site, rep_group) %>%
                         droplevels),
                   by = c("adaptedDervLabel" = "adaptedDervLabel")) %>%
  droplevels

g1 <- print(
  ggplot(data = dist_usvi_mtab.df)
  + theme_bw()
  + geom_boxplot(aes(x = Site, y = dissimilarity, 
                     # color = Site, 
                     # color = sampling_time, 
                     group = interaction(Site, sample_type)), 
                 color = "black",
                 position = position_dodge2(padding = 0.2, preserve = "single"),
                 show.legend = FALSE, outliers = FALSE)
  + geom_point(aes(x = Site, y = dissimilarity, fill = Site, group = interaction(Site, sample_type), shape = sample_type), 
               alpha = 1.0, size = 3, position = position_jitterdodge(dodge.width = 0.75, seed = 48105, jitter.width = 0.2))
  + scale_shape_manual(values = c(22, 21, 23), labels = c(sampling_time_lookup, "NA"), breaks = c(names(sampling_time_lookup), NA))
  # + scale_color_manual(values = sampling_time_colors, labels = sampling_time_lookup, breaks = names(sampling_time_lookup))
  # + scale_color_manual(values = site_colors, labels = site_lookup, breaks = names(site_lookup))
  + scale_fill_manual(values = site_colors, labels = site_lookup, breaks = names(site_lookup))
  + scale_y_continuous(expand = expansion(mult = c(0,0.1)), name = "Bray-Curtis dissimilarity")
  + scale_x_discrete(labels = site_lookup, name = "Site")
  +labs(title = "A) Targeted Metabolome")
  + theme(axis.title = element_text(size = 12, face = "bold", colour = "grey30"),
          panel.background = element_blank(), panel.border = element_rect(fill = "NA", colour = "grey30"),
          panel.grid = element_blank(),
          legend.position = "right",
          legend.key = element_blank(),
          legend.title = element_text(size = 12, face = "bold", colour = "grey30"),
          legend.text = element_text(size = 12, colour = "grey30"))
  + guides(color = "none",
           fill = guide_legend(order = 2, ncol = 1, title = "Site", direction = "vertical",
                               override.aes = list(color = "black", stroke = 1, shape = 21, size = 2)),
           shape = guide_legend(order = 1, ncol = 1, title = "Sampling time", direction = "vertical",
                                override.aes = list(color = "black", stroke = 1, size = 2)))
)

#ggsave("/Volumes/KujLab/Brianna/Projects/2022_0328_CINAR_BC/BC_CoralSeagrass_Temporal/USVI_2021_CINAR_temporal_R_Script/figures/BrayCurtis_Dissimilarity_TargetedMetabolome_20241217.png",plot = last_plot(),height=4.25,width=6.5,units = "in",dpi=600)
```

### Bray-Curtis Dissimilarity Statistical Comparison within sites across time
```{r bray curtis dissimilarity stats}

# Function to perform Kruskal-Wallis test and FDR adjustment for a given site
KW_samplingTime_dissimiliarty <- function(data, site_filter) {
  data %>%
    ungroup() %>%
    filter(Site == site_filter) %>%
    summarise(
      Kruskal_p = kruskal.test(dissimilarity ~ sample_type)$p.value,
      .groups = "drop")
}

# List of subsets with corresponding labels
subset_list <- list(
  kw_SG = list(Site = "LB_seagrass"),
  kw_TK = list(Site = "Tektite"),
  kw_YZ = list(Site = "Yawzi")
)

# Combine all results into one summary table
set.seed(123)
summary_table_3 <- purrr::imap_dfr(
  subset_list,
  ~ KW_samplingTime_dissimiliarty(dist_usvi_mtab.df, .x$Site) %>%
      mutate(Site = .x$Site) # Use .x$Site to extract the site name
)

# View the summary table
print(summary_table_3)

```
### Bray-Curtis Dissimilarity Statistical Comparison between sites within time
```{r KW and Wilcoxon test across sites within time}

# Function to perform Kruskal-Wallis test and FDR adjustment for a given sampling time
run_kruskal_time_fdr <- function(data, sample_type_filter) {
  data %>%
    ungroup() %>%
    filter(sample_type == sample_type_filter) %>%
    summarise(
      Kruskal_p = kruskal.test(dissimilarity ~ Site)$p.value,
      .groups = "drop"
    ) %>%
    mutate(
      FDR_adjusted_p = p.adjust(Kruskal_p, method = "fdr"),
      sig_05_fdr = FDR_adjusted_p < 0.05
    )
}

# Function to perform pairwise Wilcoxon rank-sum tests for a given sampling time
run_pairwise_tests <- function(data, sample_type_filter) {
  # Filter data for the specific sampling time
  filtered_data <- data %>%
    filter(sample_type == sample_type_filter)
  
  # Ensure 'dissimilarity' and 'Site' are vectors
  pairwise_result <- pairwise.wilcox.test(
    x = filtered_data$dissimilarity,
    g = filtered_data$Site,
    p.adjust.method = "fdr"
  )
  
  # Convert results to a tidy dataframe
  broom::tidy(pairwise_result) %>%
    mutate(samplingTime = sample_type_filter)
}

# List of subsets with corresponding labels
subset_list <- list(
  kw_dawn = list(samplingTime = "dawn"),
  kw_afternoon = list(samplingTime = "peak_photo")
)

# Run Kruskal-Wallis tests
summary_table_4 <- purrr::imap_dfr(
  subset_list,
  ~ run_kruskal_time_fdr(dist_usvi_mtab.df, .x$samplingTime) %>%
      mutate(samplingTime = .x$samplingTime) 
)

# Run pairwise Wilcoxon tests for significant Kruskal-Wallis results
pairwise_results <- purrr::imap_dfr(
  subset_list,
  ~ run_pairwise_tests(dist_usvi_mtab.df, .x$samplingTime)
)

# View the results
print(summary_table_4)  # Kruskal-Wallis test results
print(pairwise_results)  # Pairwise Wilcoxon test results

```

## PERMANOVA
### Categorical variables - does sampling day matter? 
```{r PERMANOVA sampling day}

permanova_dm <- temporal_wide_outliersRmv_log2%>%
  select(adaptedDervLabel, 
         Site, 
         Sampling_Day,
         sample_type,
         c("3'AMP":"valine")
         )%>%
  drop_na() # Remove rows with NaN values before the transformation

permanova.mat <- permanova_dm%>%
  select(c("3'AMP":"valine"))%>%
  as.matrix()

rownames(permanova.mat) <- permanova_dm$adaptedDervLabel

set.seed(123)

result_combined <- adonis2(
  permanova.mat ~ Site + sample_type + Sampling_Day,
  data = permanova_dm,
  method = "bray",
  permutations = 999,
  by = "terms"
)

print(result_combined)

# Extract results from the combined PERMANOVA model
results_df <- data.frame(
  Factor = rownames(result_combined),  # Automatically uses the factor names from the combined model
  F_value = result_combined$F,        # Extract F-values
  R2 = result_combined$R2,            # Extract R² values
  P_value = result_combined$`Pr(>F)`  # Extract p-values
)

print(results_df)

# Format the results_df for manuscript presentation
results_df %>%
  mutate(
    F_value = round(F_value, 2),
    R2 = round(R2, 3),
    P_value = ifelse(P_value < 0.001, "<0.001", round(P_value, 3))
  ) %>%
  kable(
    format = "html",  # Change to "latex" for LaTeX documents
    col.names = c("Factor", "F Value", "R²", "P Value"),
    caption = "Table 1. Targeted metabolome PERMANOVA results for environmental and biological variables",
    align = "l"  # Center align all columns
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "center"
  ) %>%
  column_spec(2:4, bold = TRUE) %>%
  add_header_above(c(" " = 1, "PERMANOVA Statistics" = 3))

```

## CAPSCALE model of continuous environmental & biological variables
```{r linear model}

#Log2 transform continuous variables to make them on the same general scale
cleanData <- temporal_wide_outliersRmv_log2 %>%
  select(adaptedDervLabel, Site, sample_type, NPOC, TN, Prochlorococcus, Synechococcus, Picoeukaryotes,Unpigmented_cells, c("3'AMP":valine)) %>%
  mutate(across(c(NPOC:Unpigmented_cells), ~log2(. + 1))) %>%
  drop_na() #removal of rows that have NA

#Check for co-linearity of variables
cor(cleanData[, c("Prochlorococcus", "Synechococcus", "Picoeukaryotes", "NPOC", "TN")])

cleanData.mat <- cleanData%>%
  select(c("3'AMP":"valine"))%>%
  as.matrix()

rownames(cleanData.mat) <- cleanData$adaptedDervLabel

#Calculate bray curtis dsitnace of metabolome data 
set.seed(123)
bray_dist <- vegan::vegdist(cleanData.mat, method = "bray", binary = FALSE, na.rm = TRUE)

#Run RDA using capscale
set.seed(123)
db_rda <- capscale(bray_dist ~ Site + sample_type + Prochlorococcus + Picoeukaryotes + NPOC + TN, data = cleanData)
summary(db_rda)

#Calculate significance with an ANOVA
set.seed(123)
anova(db_rda, by = "terms")



```
## Kruskal-Wallis
### Kruskal-Wallis: Within Site/Sampling Time but across days
Since metabolite samples were measured at the same three sites, at the same time of day (dawn and afternoon) every day over four days, we need to test if the metabolite concentrations within a site/sampling time (e.g., Tektite - dawn) vary over those four days. If they do, metabolites will need to be further investigate on a day-by-day basis. If not, then metabolites can be grouped by their sampling time across the four days, thus increasing our statistical power when investigating differences between sampling time. 
```{r KW across days}

# Function to perform Kruskal-Wallis test and FDR adjustment for a given subset of data
run_kruskal_fdr <- function(data, site_filter, sample_type_filter) {
  data %>%
    ungroup() %>%
    filter(Site == site_filter, sample_type == sample_type_filter) %>%
    group_by(metabolites) %>%
    summarise(
      Kruskal_p = kruskal.test(log2_concen ~ Sampling_Day)$p.value,
      .groups = "drop"
    ) %>%
    mutate(
      FDR_adjusted_p = p.adjust(Kruskal_p, method = "fdr"),
      sig_05_fdr = FDR_adjusted_p < 0.05
    )
}

# List of subsets with corresponding labels
subset_list <- list(
  kw_SG_peak = list(site = "LB_seagrass", sample_type = "peak_photo"),
  kw_SG_dawn = list(site = "LB_seagrass", sample_type = "dawn"),
  kw_TK_peak = list(site = "Tektite", sample_type = "peak_photo"),
  kw_TK_dawn = list(site = "Tektite", sample_type = "dawn"),
  kw_YZ_peak = list(site = "Yawzi", sample_type = "peak_photo"),
  kw_YZ_dawn = list(site = "Yawzi", sample_type = "dawn")
)

# Combine all results into one summary table
set.seed(123)
summary_table_5 <- purrr::imap_dfr(
  subset_list,
  ~ run_kruskal_fdr(temporal_long_wLimits_outliersRmv, .x$site, .x$sample_type) %>%
      mutate(site_sample = .y)
)

# View the summary table
print(summary_table_5)

```

#### Count number of significant metabolites
```{r count_significant_metabolites, echo=FALSE}

# Calculate the count of significant metabolites
num_significant <- summary_table_5 %>%
  filter(FDR_adjusted_p < 0.05) %>%
  nrow()

# Output the count
cat("The number of metabolites with an adjusted p-value (FDR) < 0.05 within a Site/Sampling Time but across days is:", num_significant)

```

### Kruskal-Wallis: Within Site between sampling time (dawn and afternoon)
Since metabolite concentrations did not vary significantly post-FDR correction, metabolites can be grouped by their site and sampling time and tested for signficant differences in concentration between dawn and afternoon. 

```{r KW sampling time}

# Function to perform Kruskal-Wallis test and FDR adjustment for a given site
KW_samplingTime_fdr <- function(data, site_filter) {
  data %>%
    ungroup() %>%
    filter(Site == site_filter) %>%
    group_by(metabolites) %>%
    summarise(
      Kruskal_p = kruskal.test(log2_concen ~ sample_type)$p.value,
      .groups = "drop"
    ) %>%
    mutate(
      FDR_adjusted_p = p.adjust(Kruskal_p, method = "fdr"),
      sig_05_fdr = FDR_adjusted_p < 0.05
    )
}

# List of subsets with corresponding labels
subset_list <- list(
  kw_SG = list(site = "LB_seagrass"),
  kw_TK = list(site = "Tektite"),
  kw_YZ = list(site = "Yawzi")
)

# Combine all results into one summary table
set.seed(123)
summary_table_6 <- purrr::imap_dfr(
  subset_list,
  ~ KW_samplingTime_fdr(temporal_long_wLimits_outliersRmv, .x$site) %>%
      mutate(site = .x$site) # Use .x$site to extract the site name
)

# View the summary table
print(summary_table_6)

sig_FDR_mtabs <- summary_table_6%>%
  filter(FDR_adjusted_p < 0.05)

```

#### Boxplots - significant metabolites dawn vs. afternoon
```{r plot significant mtabs}

x <- sig_FDR_mtabs%>%
  select(metabolites)%>%
  deframe() 

temporal_long_wLimits_outliersRmv%>%
  filter(metabolites %in% x)%>%
  ggplot(aes(x=Site,y=log2_concen,color=Site,linetype = sample_type),linewidth =2)+
  geom_boxplot()+
  facet_wrap(~metabolites,scales= "free")+
  scale_shape_manual(values = c(21,22))+
  scale_color_manual(values = CINAR_temporal_palette)+
  theme_bw(base_size = 12)


```

#### Heatmap - Significant metaboltites dawn vs. afternoon
```{r sig mtab heatmaps}

# Create individual heatmaps for each group

x <- sig_FDR_mtabs%>%
  filter(site == "Tektite")%>%
  select(metabolites)%>%
  deframe()

heatmap1 <- temporal_long_wLimits_outliersRmv%>%
  filter(Site == "Tektite", metabolites %in% x )%>%
  mutate(date_time = factor(date_time))%>%
  group_by(metabolites,date_time,sample_type)%>%
  mutate(median_conc = round(median(concentration),3),
         forHeatmap = if_else(median_conc < LOD, NA, median_conc))%>%
  ggplot(aes(date_time, metabolites, fill = log2(forHeatmap+1))) +
  geom_tile(color="black", aes(fill = forHeatmap)) +
  scale_fill_viridis(trans = 'log', na.value = "white", direction = 1,name = NULL,
                     labels = label_number(accuracy = 0.001))+ # Round to 2 decimal places  
  theme_minimal(base_size = 14)+
  theme(legend.position = "right", 
        text = element_text(colour = "black"),
        axis.title.x = element_blank(), 
        axis.text.x = element_blank(), 
        axis.ticks.x = element_blank(),
        axis.title.y = element_blank(),
        plot.margin = margin(10, 10, 10, 10),
        panel.border = element_rect(fill = NA, color = "black", size = 1),
        panel.grid = element_blank())+         # Remove grid lines
  ggtitle("Tektite")

x <- sig_FDR_mtabs%>%
  filter(site == "Yawzi")%>%
  select(metabolites)%>%
  deframe()

heatmap2 <- temporal_long_wLimits_outliersRmv%>%
  filter(Site == "Yawzi", metabolites %in% x )%>%
  mutate(date_time = factor(date_time))%>%
  group_by(metabolites,date_time,sample_type)%>%
  mutate(median_conc = round(median(concentration),3),
         forHeatmap = if_else(median_conc < LOD, NA, median_conc))%>%
  ggplot(aes(date_time, metabolites, fill = log2(forHeatmap+1))) +
  geom_tile(color="black", aes(fill = forHeatmap)) +
  scale_fill_viridis(trans = 'log', na.value = "white", direction = 1,name = NULL,
                     labels = label_number(accuracy = 0.001))+ # Round to 2 decimal places  
  theme_minimal(base_size = 14)+
    theme(legend.position = "right", 
        text = element_text(colour = "black"),
        axis.title.x = element_blank(), 
        axis.text.x = element_blank(), 
        axis.ticks.x = element_blank(),
        axis.title.y = element_blank(),
        plot.margin = margin(10, 10, 10, 10),
        panel.border = element_rect(fill = NA, color = "black", size = 1),
        panel.grid = element_blank())+         # Remove grid lines
  ggtitle("Yawzi")

x <- sig_FDR_mtabs%>%
  filter(site == "LB_seagrass")%>%
  select(metabolites)%>%
  deframe()

heatmap3 <- temporal_long_wLimits_outliersRmv%>%
  filter(Site == "LB_seagrass", metabolites %in% x )%>%
  mutate(date_time = factor(date_time))%>%
  group_by(metabolites,date_time,sample_type)%>%
  mutate(median_conc = round(median(concentration),3),
         forHeatmap = if_else(median_conc < LOD, NA, median_conc))%>%
  ggplot(aes(date_time, metabolites, fill = log2(forHeatmap+1))) +
  geom_tile(color="black", aes(fill = forHeatmap)) +
  scale_fill_viridis(trans = 'log', na.value = "white", direction = 1,name = NULL,
                     labels = label_number(accuracy = 0.001))+ # Round to 2 decimal places  
  theme_minimal(base_size = 14)+
  labs(x="Sampling Day")+
      theme(legend.position = "right", 
        text = element_text(colour = "black"),
        axis.title.y = element_blank(),
        plot.margin = margin(10, 10, 10, 10),
        panel.border = element_rect(fill = NA, color = "black", size = 1),
        panel.grid = element_blank())+         # Remove grid lines
  scale_x_discrete(labels=c('Day2','Day2',
                            'Day3','Day3',
                            'Day4','Day4',
                            'Day5','Day5'))+
 ggtitle("Lameshur Bay Seagrass")


# Create the small dawn/afternoon heatmap
dawn_heatmap <- temporal_long_wLimits_outliersRmv%>%
    mutate(date_time = factor(date_time))%>%
  ggplot(aes(x = date_time, y = 1, fill = sample_type)) +
  geom_tile() +
  scale_fill_manual(values = c("dawn" = "black", "peak_photo" = "white")) +
  theme_bw() + # Remove all grid, text, and axes
    theme(legend.position = "none", 
        text = element_text(colour = "black"),
        axis.title.x = element_blank(), 
        axis.text.x = element_blank(), 
        axis.ticks.x = element_blank(),
        axis.title.y = element_blank(), 
        axis.text.y = element_blank(), 
        axis.ticks.y = element_blank(),
        plot.margin = margin(2, 2, 2, 2),
        panel.border = element_rect(fill = NA, color = "black", size = 1),
        panel.grid = element_blank())         # Remove grid lines


# Combine the heatmaps with white space in between
combined_plot <- heatmap1 / heatmap2 / heatmap3 / dawn_heatmap+ 
  plot_layout(ncol = 1, heights = c(.1875, 0.375, .75,0.03)) & 
  plot_annotation(theme = theme(plot.background = element_rect(fill = "white", color = NA)))

# Show the final plot
combined_plot

#ggsave("Z:/Brianna/Projects/2022_0328_CINAR_BC/BC_CoralSeagrass_Temporal/USVI_2021_CINAR_temporal_R_Script/figures/Temporally_Significant_TargetedMetabolites_log10_concentration_20250204.png",plot = last_plot(),height=8.5,width=8,units = "in",dpi=600)

```

### Kruskal-Wallis: Between sites within a sampling timepoint

```{r KW site}

# Function to perform Kruskal-Wallis test and FDR adjustment for a given site
KW_site<- function(data, time_filter) {
  data %>%
    ungroup() %>%
    filter(sample_type == time_filter) %>%
    group_by(metabolites) %>%
    summarise(
      Kruskal_p = kruskal.test(log2_concen ~ Site)$p.value,
      .groups = "drop"
      )
}

# List of subsets with corresponding labels
subset_list <- list(
  kw_dawn = list(sample_type = "dawn"),
  kw_peak = list(sample_type = "peak_photo")
)

# Combine all results into one summary table
set.seed(123)
summary_table_7 <- purrr::imap_dfr(
  subset_list,
  ~ KW_site(temporal_long_wLimits_outliersRmv, .x$sample_type) %>%
      mutate(sample_type = .x$sample_type) # Use .x$sample_type to extract the sample_type name
)

# View the summary table
print(summary_table_7)

sig_KW_mtabs <- summary_table_7%>%
  filter(Kruskal_p < 0.05)

```
### Pairwise Wilcoxon Rank Sum Test
#### Dawn
```{r site wilcoxon rank sum test}

# Filter significant metabolites at dawn
significant_metabolites <- sig_KW_mtabs %>%
  filter(sample_type == "dawn") %>%
  pull(metabolites)

# Subset the original data for significant metabolites and "dawn" sample type
dawn_data <- temporal_long_wLimits_outliersRmv %>%
  filter(sample_type == "dawn", metabolites %in% significant_metabolites)

# Perform pairwise Wilcoxon tests for each significant metabolite

set.seed(123)
pairwise_results <- dawn_data %>%
  group_by(metabolites) %>%
  summarise(
    pairwise_wilcox = list(
      pairwise.wilcox.test(
        x = log2_concen,
        g = Site,
        p.adjust.method = "fdr"
      )
    ),
    .groups = "drop"
  )

# View results
pairwise_results <- pairwise_results %>%
  mutate(pairwise_summary = map(pairwise_wilcox, ~ broom::tidy(.x)))

# Optionally, flatten for easier inspection
pairwise_summary <- pairwise_results %>%
  unnest(pairwise_summary)%>%
  filter(p.value < 0.05)%>%
  group_by(metabolites)%>%
  mutate(n = n(),
         sample_type = "dawn")

wilcoxon_Site_dawn_summary <- pairwise_summary

# Print results
print(wilcoxon_Site_dawn_summary)

paste(length(unique(wilcoxon_Site_dawn_summary$metabolites)),"metabolites were found significant between sites at Dawn")

sig_dawn_site_mtabNames <- unique(wilcoxon_Site_dawn_summary$metabolites)
```

##### Boxplots - Wilcoxon - Dawn
```{r plot significant mtabs}

x <- sig_dawn_site_mtabNames

bp_wilcoxon_by_site_dawn <- temporal_long_wLimits_outliersRmv%>%
  filter(metabolites %in% x, sample_type == "dawn")%>%
  ggplot(aes(x=Site,y=log2_concen,color=Site),linewidth =2)+
  geom_boxplot()+
  facet_wrap(~metabolites,scales= "free")+
  scale_shape_manual(values = c(21,22))+
  scale_color_manual(values = CINAR_temporal_palette)+
  theme_bw(base_size = 12)+
  labs(title = "Significant (FDR Adj. p-value < 0.05) by Site (Dawn)")

print(bp_wilcoxon_by_site_dawn)

```
#### Afternoon
```{r site wilcoxon rank sum test}

# Filter significant metabolites at peak_photo
significant_metabolites <- sig_KW_mtabs %>%
  filter(sample_type == "peak_photo") %>%
  pull(metabolites)

# Subset the original data for significant metabolites and "peak_photo" sample type
dawn_data <- temporal_long_wLimits_outliersRmv %>%
  filter(sample_type == "peak_photo", metabolites %in% significant_metabolites)

# Perform pairwise Wilcoxon tests for each significant metabolite
set.seed(123)
pairwise_results <- dawn_data %>%
  group_by(metabolites) %>%
  summarise(
    pairwise_wilcox = list(
      pairwise.wilcox.test(
        x = log2_concen,
        g = Site,
        p.adjust.method = "fdr"
      )
    ),
    .groups = "drop"
  )

# View results
pairwise_results <- pairwise_results %>%
  mutate(pairwise_summary = map(pairwise_wilcox, ~ broom::tidy(.x)))

# Optionally, flatten for easier inspection
pairwise_summary <- pairwise_results %>%
  unnest(pairwise_summary)%>%
  filter(p.value < 0.05)%>%
  group_by(metabolites)%>%
  mutate(n = n(),
         sample_type = "afternoon")

wilcoxon_Site_peak_summary <- pairwise_summary

# Print results
print(wilcoxon_Site_peak_summary)

paste(length(unique(wilcoxon_Site_peak_summary$metabolites)),"metabolites were found significant between sites at Dawn")

sig_peak_site_mtabNames <- unique(wilcoxon_Site_peak_summary$metabolites)
```

##### Boxplots - Wilcoxon - Afternoon
```{r plot significant mtabs}

x <- sig_peak_site_mtabNames 

bp_wilcoxon_by_site_peak <- temporal_long_wLimits_outliersRmv%>%
  filter(metabolites %in% x, sample_type == "peak_photo")%>%
  ggplot(aes(x=Site,y=log2_concen,color=Site),linewidth =2)+
  geom_boxplot()+
  facet_wrap(~metabolites,scales= "free")+
  scale_shape_manual(values = c(21,22))+
  scale_color_manual(values = CINAR_temporal_palette)+
  theme_bw(base_size = 12)+
  labs(title = "Significant (FDR Adj. p-value < 0.05) by Site (Afternoon)")

print(bp_wilcoxon_by_site_peak)

```
#### Commonalities and Differences
```{r Wilcoxon dawn afternoon comparison}

# Join summary tables 

wilcoxon_Site_summary <- rbind(wilcoxon_Site_dawn_summary,wilcoxon_Site_peak_summary)%>%
  mutate(pairs = paste0(group1,"_",group2))

# Join both vectors and remove duplicates
unique_combined <- unique(c(sig_peak_site_mtabNames, sig_dawn_site_mtabNames))

# Find common and distinct values
common_values <- intersect(sig_peak_site_mtabNames, sig_dawn_site_mtabNames)
distinct_values <- c(setdiff(sig_peak_site_mtabNames, sig_dawn_site_mtabNames),
                     setdiff(sig_dawn_site_mtabNames, sig_peak_site_mtabNames))

# Verify totals
length(unique_combined) == length(common_values) + length(unique(distinct_values))

```
##### Plot common metabolites (facet sample_type)
```{r plot common by site}

# Define the output PDF file
#output_pdf <- "Z:/Brianna/Projects/2022_0328_CINAR_BC/BC_CoralSeagrass_Temporal/USVI_2021_CINAR_temporal_R_Script/temp_outputs/Wilcoxon_Sig_FDR05_Mtabs_bySite_Common_SampleTime_20250109.pdf"

# Open the PDF device
#pdf(output_pdf, width = 11, height = 8.5)

# Loop through each metabolite and generate the plot
for (metabolite in common_values) {
  # Filter data for the current metabolite
  plot_data <- temporal_long_wLimits_outliersRmv %>%
    filter(metabolites == metabolite)
  
  # Generate the plot
  p <- ggplot(plot_data, aes(x = Site, y = log2_concen, color = Site), linewidth = 2) +
    geom_boxplot() +
    facet_wrap(~sample_type) +
    scale_shape_manual(values = c(21, 22)) +
    scale_color_manual(values = CINAR_temporal_palette) +
    theme_bw(base_size = 12) +
    labs(
      title = paste("Significant (FDR Adj. p-value < 0.05) by Site"),
      subtitle = metabolite
    )
  
  # Print the plot to the PDF device
  print(p)
}

# Close the PDF device
dev.off()

# Notify the user
#cat("Plots saved to", output_pdf, "\n")

```
##### Plot unique metabolites (facet sample_type)
```{r plot significant mtabs}

# Define the output PDF file
#output_pdf <- "Z:/Brianna/Projects/2022_0328_CINAR_BC/BC_CoralSeagrass_Temporal/USVI_2021_CINAR_temporal_R_Script/temp_outputs/Wilcoxon_Sig_FDR05_Mtabs_bySite_Unique_SampleTime_20250109.pdf"

# Open the PDF device
#pdf(output_pdf, width = 11, height = 8.5)

# Loop through each metabolite and generate the plot
for (metabolite in distinct_values) {
  # Filter data for the current metabolite
  plot_data <- temporal_long_wLimits_outliersRmv %>%
    filter(metabolites == metabolite)
  
  # Generate the plot
  p <- ggplot(plot_data, aes(x = Site, y = log2_concen, color = Site), linewidth = 2) +
    geom_boxplot() +
    facet_wrap(~sample_type) +
    scale_shape_manual(values = c(21, 22)) +
    scale_color_manual(values = CINAR_temporal_palette) +
    theme_bw(base_size = 12) +
    labs(
      title = paste("Significant (FDR Adj. p-value < 0.05) by Site"),
      subtitle = metabolite
    )
  
  # Print the plot to the PDF device
  print(p)
}

# Close the PDF device
dev.off()

# Notify the user
#cat("Plots saved to", output_pdf, "\n")

```
# Manuscript figures
## Figure 1. Map of St. John USVI
```{r figure 1}

# Raw shapefiles
usa <- st_read("./Analysis/R/scripts/USVI_map/stanford-vt021tk4894-shapefile/","vt021tk4894")
sttstj <- st_read("./Analysis/R/scripts/USVI_map/stsj_fin")
nps <- st_read("./Analysis/R/scripts/USVI_map/NPS_-_Land_Resources_Division_Boundary_and_Tract_Data_Service")

# site metadata

temporal_sites <- read.table("./BC_CoralSeagrass_Temporal/Documents/temporal_sites_updated20250130.txt", sep = "\t", header = TRUE)
temporal_sites$Site <- factor(temporal_sites$Site, levels = c( "Yawzi", "Tektite","LB_seagrass"))
usvi <- usa %>% filter(state == "United States Virgin Islands")

reefZONE <- sttstj %>%
  filter(ZONE %in% c("Forereef", "Reef Crest", "Backreef"))

vinp <- nps %>% filter(PARKNAME == "Virgin Islands")

#plot
temporal_map <- ggplot() +
  geom_sf(data = usvi, fill = "#e5dccd",color = "#6699CC") +
  geom_sf(data = reefZONE, fill = "#cc6677", color = "#882255",alpha=0.5) +
  geom_sf(data = vinp, fill = NA, color = "#117733", linewidth = 0.6) +
  coord_sf(xlim = c(-64.8040, -64.655), ylim = c(18.2895, 18.378364), expand = FALSE) +
  geom_point(data = temporal_sites, mapping = aes(x = Lon, y = Lat, fill = Site),colour="#333333", pch = 21, size = 3) +
  scale_fill_manual(values = site_colors, labels = site_lookup, breaks = names(site_lookup))+
  annotation_scale(location = "bl", width_hint = 0.2) +
  theme(panel.grid.major = element_blank(), panel.background = element_rect(fill = "#6699CC"),legend.position = c(0.902, 0.21))+
  labs(x = "Longitude",y="Latitude")+
  geom_text() +
  annotate("text", label = "St. John,\nU.S. Virgin Islands",x = -64.74, y = 18.34, size = 4, colour = "black",fontface=2)+
  annotate("text", label = "Coral reef habitat",x = -64.785, y = 18.307, size = 3, colour = "#882255",fontface=2)+
  annotate("text", label = "Virgin Islands \nNational Park",x = -64.715, y = 18.3585, size = 3, colour = "#117733",fontface=2)+
  annotate("text", label = "Lameshur Bay",x = -64.735, y = 18.305, size = 3, colour = "black",fontface=2)

temporal_map + geom_segment(aes(x = -64.735, y = 18.307, xend = -64.728, yend = 18.312),
                            arrow = arrow(length = unit(0.01, "npc")))

```

```{r figure 1 zoom}

#plot
temporal_map_zoom <- ggplot() +
  geom_sf(data = usvi, fill = "#e5dccd",color = "#6699CC") +
  geom_sf(data = reefZONE, fill = "#cc6677", color = "#882255",alpha=0.5) +
  geom_sf(data = vinp, fill = NA, color = "#117733", linewidth = 0.6) +
  coord_sf(xlim = c(-64.727, -64.72), ylim = c(18.308, 18.320), expand = FALSE) +
  geom_point(data = temporal_sites, mapping = aes(x = Lon, y = Lat, fill = Site),colour="#333333", pch = 21, size = 5,alpha=0.88) +
  scale_fill_manual(values = site_colors, labels = site_lookup, breaks = names(site_lookup))+
  annotation_scale(location = "bl", width_hint = 0.2) +
  theme(panel.grid.major = element_blank(), panel.background = element_rect(fill = "#6699CC"),legend.position = "top",
  axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
  labs(x = "Longitude",y="Latitude")+
  geom_text() +
  annotate("text", label = "Great Lameshur Bay",x = -64.723, y = 18.315, size = 2.5, colour = "black",fontface=2)

temporal_map_zoom

```

```{r save figure 1}

#save
#png(file = "./BC_CoralSeagrass_Temporal/figures/CINAR_temporal_USVI_StJohn_Sampling_Map.png",width = 8.5, height = 7, units = "in", res = 300)

#save
#png(file = "./BC_CoralSeagrass_Temporal/figures/CINAR_temporal_USVI_StJohn_Sampling_Map_zoom.png",width = 6, height = 5.5, units = "in", res = 300)

```

```{r removal map variables}
# 
# rm(vinp,usvi,usa,sttstj,temporal_sites,reefZONE,nps)

```

## Figure 2. All Quantified Mtab Heatmap
```{r figure 2}

fig_2 <- temporal_long_wLimits_outliersRmv %>%
  ungroup()%>%
  mutate(concen_forFigure = case_when(concentration < LOD ~ NA, TRUE ~ concentration))%>%
  select(metabolites, Site, sample_type,concen_forFigure, LOD, LOQ)%>%
  mutate(sample_type = case_when(sample_type == "peak_photo" ~ "afternoon", 
                                 TRUE ~ sample_type),
  `Site:Sampling Period` = paste(Site, sample_type)) %>%
  group_by(metabolites,`Site:Sampling Period`)%>%
  mutate(median_conc = median(concen_forFigure,na.rm = TRUE),
         `Site:Sampling Period` = factor(`Site:Sampling Period`, levels = c("LB_seagrass dawn", "LB_seagrass afternoon", "Tektite dawn", "Tektite afternoon", "Yawzi dawn", "Yawzi afternoon")))%>%
  ggplot(aes(x = `Site:Sampling Period`, y = metabolites)) +
  geom_tile(color="black",aes(fill = median_conc)) +
  scale_fill_viridis(trans = 'log', na.value = "white", direction = 1,
                     breaks = c(.03,0.3,3), "Median Concentration (nM)") + 
  theme_bw(base_size = 14)+
  theme(legend.position = "top",text = element_text(colour = "black"))+
  labs(x = "Site", y="Metabolite")

print(fig_2)

#ggsave("Z:/Brianna/Projects/2022_0328_CINAR_BC/BC_CoralSeagrass_Temporal/USVI_2021_CINAR_temporal_R_Script/figures/FigureS1_Mtab_Heatmap_medianConcentration_nM_v2_20250204.pdf",plot = fig_2,height = 11,width = 8.5,dpi = 300)


```

## Figure 3. 
```{r figure 3}

```
## Figure 4. 
```{r figure 4}

```
## Figure 4.
```{r figure 5}

```

# Plots for presentations 
##Temporal metabolites - over time
```{r} 

selected_mtab = "taurine"

temporal_long_wLimits_outliersRmv %>%
  filter(metabolites == selected_mtab) %>%
   mutate(sample_type = case_when(sample_type == "peak_photo" ~ "afternoon", 
                                 TRUE ~ sample_type),
          day_time = paste(Sampling_Day,sample_type),
          day_time = factor(day_time,levels = c("Day2 dawn", "Day2 afternoon", 
                                                "Day3 dawn", "Day3 afternoon",
                                                "Day4 dawn", "Day4 afternoon",
                                                "Day5 dawn", "Day5 afternoon")))%>%

  mutate(Site = factor(Site, levels = c("Tektite","Yawzi","LB_seagrass")))%>%
  group_by(Site, day_time) %>%
  summarize(
    mean_concentration = mean(concentration, na.rm = TRUE),
    sd_concentration = sd(concentration, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = day_time, y = mean_concentration, fill = Site)) + 
  geom_errorbar(aes(ymin = mean_concentration - sd_concentration,
                    ymax = mean_concentration + sd_concentration), 
                width = 0.2) +
  geom_point(size = 3, shape = 21, color = "black") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position="bottom") +
  scale_fill_manual(values = site_colors, labels = site_lookup, breaks = names(site_lookup))+
  scale_color_manual(values = site_colors, labels = site_lookup, breaks = names(site_lookup))+
  facet_wrap(~Site, ncol=1, scales = "free")+
  labs(title = selected_mtab)

```

# Manuscript tables

## Table S1 - Water Chemistry numbers and statistics 
```{r table s1}

table_s1 <- temporal_wide_outliersRmv_log2 %>%
  select(Site,sample_type,NPOC,TN,Prochlorococcus,Synechococcus,Picoeukaryotes,Unpigmented_cells)%>%
  mutate(sample_type = case_when(sample_type == "peak_photo" ~ "afternoon", 
                                 TRUE ~ sample_type),
         `Site:Sampling Period` = paste(Site, sample_type)) %>%
  select(-c(Site, sample_type)) %>%
  pivot_longer(cols = c(NPOC,TN,Prochlorococcus,Synechococcus,Picoeukaryotes,Unpigmented_cells), 
               names_to = "Measurement", 
               values_to = "Amount")%>%
  group_by(`Site:Sampling Period`, Measurement) %>%
  summarise(mean = mean(Amount, na.rm = TRUE), 
            sdev = sd(Amount, na.rm = TRUE))%>%
  mutate(mean_dev = paste0(round(mean, 2), " ± ", round(sdev, 2)))%>%
  mutate(Analysis = case_when(Measurement %in% c("NPOC","TN") ~ "Nutrients",
                                 TRUE ~ "FCM"),
         Units = case_when(Measurement %in% c("NPOC","TN") ~ "uM",
                                 TRUE ~ "counts"))%>%
  select(Analysis,Units,`Site:Sampling Period`, mean_dev,Measurement)%>%
  pivot_wider(names_from = `Site:Sampling Period`, values_from = mean_dev)%>%
  arrange(desc(Analysis))

#write.csv(table_s1, "Z:/Brianna/Projects/2022_0328_CINAR_BC/BC_CoralSeagrass_Temporal/USVI_2021_CINAR_temporal_R_Script/csv outputs/TableS1_Nutrients_FCM_Table_byGroup.csv")

```

## Table S2 - Quantified / Detected Metabolites
```{r table s2}


table_s2 <- temporal_long_wLimits_outliersRmv %>%
  ungroup()%>%
  mutate(concen_forTable = case_when(concentration < LOD ~ NA, TRUE ~ concentration))%>%
  select(metabolites, Site, sample_type,concen_forTable, LOD, LOQ)%>%
  mutate(sample_type = case_when(sample_type == "peak_photo" ~ "afternoon", 
                                 TRUE ~ sample_type),
         `Site:Sampling Period` = paste(Site, sample_type)) %>%
  select(-c(Site, sample_type)) %>%
  group_by(metabolites, `Site:Sampling Period`) %>%
  summarise(
    LOD = mean(LOD, na.rm = TRUE), # Use mean or another method to summarize LOD
    LOQ = mean(LOQ, na.rm = TRUE), # Use mean or another method to summarize LOQ
    mean_nM = mean(concen_forTable, na.rm = TRUE),
    sd_nM = sd(concen_forTable, na.rm = TRUE),
    concen_nM = paste0(round(mean_nM, 3), " ± ", round(sd_nM, 3)),
    .groups = "drop"
  ) %>%
  select(-c(mean_nM, sd_nM)) %>%
  pivot_wider(
    names_from = `Site:Sampling Period`,
    values_from = concen_nM
  )%>%
  mutate(LOD = round(LOD,3), 
         LOQ = round(LOQ,3))%>%
  rename(`LOD (nM)` = LOD, 
         `LOQ (nM)` = LOQ,
         `Metabolite name` = metabolites)%>%
  rowwise() %>%
  mutate(
    `n groups above LOD` = sum(!str_detect(c_across(starts_with("LB_") | starts_with("Tektite") | starts_with("Yawzi")), "NA"))
  ) %>%
  ungroup()

#need to fix this, the plus +/- are messing everything up agian

#write.csv(table_s2, "Z:/Brianna/Projects/2022_0328_CINAR_BC/BC_CoralSeagrass_Temporal/USVI_2021_CINAR_temporal_R_Script/csv outputs/TableS2_Metabolite_Quantification_Table_byGroup.csv")

```