---
title: "USVI 2021 CINAR Temporal Metabolite Analysis"
author: "Brianna Garcia"
date: "`r Sys.Date()`"
output: html_document
---

# Document Set-up
## Set working directory
```{r setup, include=FALSE}

#Work PC directories
#knitr::opts_knit$set(root.dir = "Z:/Brianna/Projects/2022_0328_CINAR_BC")
#setwd("Z:/Brianna/Projects/2022_0328_CINAR_BC")

#Laptop directories
knitr::opts_knit$set(root.dir = "/Volumes/KujLab/Brianna/Projects/2022_0328_CINAR_BC")
setwd("/Volumes/KujLab/Brianna/Projects/2022_0328_CINAR_BC")

```

## Load required packages
```{r packages, message=FALSE}

library(RColorBrewer)
library(gridExtra)
library(sf)
library(ggspatial)
library(scales)
library(data.table)
library(ggdendro)
library(viridis)
library(vegan)
library(ggpubr)
library(tidyverse)
library(lubridate)
library(knitr)
library(kableExtra)
library(purrr)
library(patchwork)

```

## Define color palette for all downstream plots
```{r color palette}

CINAR_temporal_palette=c("#A1CAF1","#AE2337","#BFB386")

if(!exists("site_lookup", envir = .GlobalEnv)){
  site_lookup <- data.frame(site = c("LB_seagrass", "Tektite", "Yawzi", "control_extraction", "control_pcr", "control_seq"),
                            label = c("Lameshur Bay seagrass", "Tektite Reef", "Yawzi Reef",
                                      "Control (DNA Extraction)", "Control (PCR)", "Control (Sequencing)")) %>%
    dplyr::mutate(label = stringr::str_wrap(label, width = 16)) %>%
    tibble::deframe(.)
  site_colors <- pals::kelly(22)[6:(5+length(site_lookup))] %>%
    # site_colors <- viridisLite::cividis(n = length(site_lookup), direction = 1) %>%
    setNames(., names(site_lookup))
  sampling_time_lookup <- data.frame(sampling_time = c("dawn", "peak_photo"),
                                     label = c("Dawn", "Afternoon")) %>%
    tibble::deframe(.)
  sampling_time_colors <- pals::ocean.haline(n = length(sampling_time_lookup)) %>%
    setNames(., names(sampling_time_lookup))
  sampling_day_lookup <- data.frame(sampling_day = c("Day1", "Day2", "Day3", "Day4", "Day5"),
                                    label = c("20210122", "20210123", "20210124", "20210125", "20210126")) %>%
    tibble::deframe(.)
  sampling_day_colors <- pals::ocean.thermal(n = length(sampling_day_lookup)) %>%
    setNames(., names(sampling_day_lookup))
}


```

# Load in data and prep for statistical analysis

## Load data
```{r load data, message=FALSE}

data <- read_csv('./Analysis/MATLAB/BC_qaqc/20230605/temporal_mergedTable_metCols_CINARBC_20230605_BMG.csv')
data <- as_tibble(data,rownames="rowID_full")
data$rep_group<-as.factor(data$rep_group)
limits <- read_csv('./Analysis/MATLAB/BC_qaqc/20230605/CINAR_metsFiltered_manualCheck_LODLOQ.2023.06.05.csv')
seq <- read_csv("Raw_data/sequence/mtab_CINAR_BC_032822_MATLABtrunc.csv")

```

## Merge in filenames
```{r merge filenames}

seq_pos <- seq%>%
  filter(ionMode == "pos", goodData == 1, sType == "rep" )%>%
  select(File_Name,sMatlabName)%>%
  rename(File_Name_Pos = File_Name)

seq_neg <- seq%>%
  filter(ionMode == "neg", goodData == 1, sType == "rep" )%>%
  select(File_Name,sMatlabName)%>%
  rename(File_Name_Neg = File_Name)

seq_mode_sidebyside <- merge(seq_neg,seq_pos,by = "sMatlabName")%>%
  rename(adaptedDervLabel = sMatlabName)

data_wFileName <- merge(seq_mode_sidebyside,data,by = "adaptedDervLabel",all.y = TRUE)

rm(seq_pos,seq_neg, seq_mode_sidebyside,data)

```

## Subset temporal data for larger dataset
```{r subset temporal data}

temporal <- data_wFileName %>%
  ungroup()%>%
  add_count(rep_group)%>%
  relocate(n,.after=rep_group)%>%
  filter(n > 1)%>%
  mutate(group = case_when(Site %in% c("Tektite","Yawzi") ~ "coral", 
                           Site == "LB_seagrass" ~ "seagrass"))%>%
  relocate(group, .after = Site)%>%
  rename(NPOC = NPOC_uM_, TN = TN_uM_)%>%
  mutate(Site = factor(Site, levels = c("LB_seagrass","Tektite","Yawzi")))%>%
  mutate(Site_abv =
           case_when(Site =="Yawzi" ~ "YZ",
                     Site =="Tektite" ~ "TK",
                     Site =="LB_seagrass" ~ "SG"),.after = Site)%>%
  mutate(Site_abv = factor(Site_abv, levels = c("YZ", "TK","SG")))%>%
  mutate(Date = parse_date_time(Date, orders = "mdy"),
         date_time = ymd_hms(paste(Date, Time)))

temporal_long <- temporal %>%
  pivot_longer(cols="2'deoxycytidine":"valine",names_to="metabolites",values_to="concentration",values_drop_na=TRUE)

limits <- limits%>%
  rename(metabolites = metNames_filtered)

temporal_long_wLimits <- merge(temporal_long,limits,by="metabolites",all.y = TRUE)

rm(temporal_long)

temporal.mat <- temporal
rownames(temporal.mat) <- temporal$adaptedDervLabel
temporal.mat <- temporal.mat[,47:102]
temporal.mat <- temporal.mat[, ! apply(temporal.mat , 2 , function(x) 
  sd(x, na.rm = TRUE)==0 ) ] #remove metabolites with a SD of zero 

x <- colnames(temporal.mat) #pull out metabolite names

temporal_long_wLimits <- temporal_long_wLimits%>%
  ungroup()%>%
  filter(metabolites %in% x,
         !is.na(LOD_filtered_nM))%>%
  mutate(QuantFlag = concentration < LOQ_filtered_nM,
         LODflag = concentration < LOD_filtered_nM)%>%
  rename(LOD = LOD_filtered_nM, LOQ = LOQ_filtered_nM)%>%
  mutate(site_time = paste0(Site,"_",sample_type))

x <- unique(temporal_long_wLimits$metabolites)

temporal_wide <- temporal_long_wLimits %>%
  select(-c(LOD,LOQ,LODflag,QuantFlag))%>%
  pivot_wider(names_from = metabolites,values_from = concentration)

temporal.mat <- as.matrix(temporal_wide[,49:97])
rownames(temporal.mat) <- temporal_wide$adaptedDervLabel

rm(limits,data_wFileName,temporal)

```
# Outlier detection
## Calculate Euclidean distances by replicate group
```{r outlier distances}

# Create a plotting area for boxplots
par(mfrow = c(2, 3))  # Adjust rows/columns to fit all groups
par(mar = c(10, 4, 4, 2))  # Bottom, Left, Top, Right

# Iterate through each site_time group
unique_site_times <- unique(temporal_long_wLimits$site_time)

for (site in unique_site_times) {
  # Filter data for the current site_time
  site_data <- temporal_long_wLimits %>%
    filter(site_time == site) %>%
    select(metabolites,concentration,adaptedDervLabel,site_time)%>%
    pivot_wider(names_from = metabolites, values_from = concentration)%>%
    select(-site_time)

  # Convert to matrix for distance calculation
  site_matrix <- as.matrix(site_data[,-1])
  rownames(site_matrix) <- site_data$adaptedDervLabel
  
  # Compute pairwise Euclidean distances
  pairwise_distances <- dist(site_matrix, method = "euclidean")
  distance_matrix <- as.matrix(pairwise_distances)  # Convert to square form
  
  # Calculate mean distances for each sample
  mean_distances <- rowMeans(distance_matrix)
  
  # Determine thresholds for outliers using IQR
  iqr_distances <- IQR(mean_distances, na.rm = TRUE)
  lower_threshold <- quantile(mean_distances, 0.25, na.rm = TRUE) - 1.5 * iqr_distances
  upper_threshold <- quantile(mean_distances, 0.75, na.rm = TRUE) + 1.5 * iqr_distances
  
  # Plot boxplot for the distance matrix
  boxplot(
    distance_matrix, 
    main = paste("Site-Time:", site), 
    ylab = "Euclidean Distance", 
    col = "lightgray", 
    outline = FALSE, 
    las = 2
  )
  
  # Add points for mean distances
  points(mean_distances, col = "blue", pch = 21, size= 3, bg = "green")
  
  # Add lines for thresholds
  abline(h = c(lower_threshold, upper_threshold), col = "red", lty = 2, lwd = 2)
}

rm(unique_site_times,distance_matrix,site_data,site_matrix,iqr_distances,
   lower_threshold,upper_threshold,mean_distances,pairwise_distances,
   site)

```

## Outlier detection using NMDS with 95% CI
```{r outlier NMDS}

# Output plot list
plots <- list()

# Loop through each site
for (site in unique(temporal_wide$Site)) {
  # Filter data for the current site
  site_meta <- temporal_wide %>% filter(Site == site)
  site_data <- temporal.mat[rownames(temporal.mat) %in% site_meta$adaptedDervLabel, ]
  
  # Run NMDS
  nmds <- metaMDS(site_data, k = 2, trymax = 100, distance = "bray")
  
  # Extract scores
  scores <- as.data.frame(scores(nmds, display = "sites"))
  scores <- cbind(scores, site_meta)
 
  # Create the plot
  library(ggrepel)
  
  p <- ggscatter(scores, x = "NMDS1", y = "NMDS2", color = "sample_type", size = 3) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_vline(xintercept = 0, linetype = 2) +
    ggtitle(paste("NMDS for Site:", site)) +
    theme_bw(base_size = 12) +
    # Add 95% confidence ellipses
    stat_ellipse(aes(color = sample_type, fill = sample_type), level = 0.95, geom = "polygon", alpha = 0.05, show.legend = FALSE) +
    # Add labels
    geom_text(aes(label = adaptedDervLabel), vjust = -0.5, size = 3)
  
  
  # Save the plot
  plots[[site]] <- p
}

# View plots
#plots[[3]]  # Example for the first site

# Save all plots
#pdf(file= "Z:/Brianna/Projects/2022_0328_CINAR_BC/BC_CoralSeagrass_Temporal/figures/CINAR_USVI2021_NMDS_by_site_wEllipses.pdf",width = 11, height = 8.5,onefile=TRUE)
for (plot in plots) print(plot)
#dev.off()

```
## Remove outliers from dataset 
```{r remove outliers}

temporal_long_wLimits_outliersRmv <- temporal_long_wLimits%>%
  ungroup()%>%
  filter(adaptedDervLabel!= "CINAR_BC_73")%>%
  mutate(log2_concen = log2(concentration+1))

temporal_wide_outliersRmv_log2 <- temporal_long_wLimits_outliersRmv%>%
  select(-c(concentration,QuantFlag,LODflag,LOD,LOQ))%>%
  pivot_wider(names_from = "metabolites",values_from = "log2_concen")

temporal.mat_outliersRmv_log2 <- temporal_wide_outliersRmv_log2%>%
  select(c("2'deoxycytidine":"valine"))%>%
  as.matrix()

rownames(temporal.mat_outliersRmv_log2) <- temporal_wide_outliersRmv_log2$adaptedDervLabel

```

## Save table with outliers removed
```{r save outlier rmvd table}

#write.csv(temporal_long_wLimits_outliersRmv,file = "/Volumes/KujLab/Brianna/Projects/2022_0328_CINAR_BC/BC_CoralSeagrass_Temporal/USVI_2021_CINAR_temporal_R_Script/csv outputs/USVI2021_CINARtemporal_BzCl_Exometabolite_QCd_longFormat_outliersRmvd_wMetadata.csv")

#write.csv(temporal.mat_outliersRmv_log2,file = "/Volumes/KujLab/Brianna/Projects/2022_0328_CINAR_BC/BC_CoralSeagrass_Temporal/USVI_2021_CINAR_temporal_R_Script/csv outputs/USVI2021_CINARtemporal_BzCl_Exometabolite_QCd_wideFormat_outliersRmvd_log2_noMetadata.csv")

```
#Data summary

## HOBO Logger (Light and Temperature)
Note: Why does Yawzi reef have such lower lumens in the afternoon? Sampling order appears to be TK, YZ, and SG.
```{r HOBO}

temporal_long_wLimits_outliersRmv%>%
  group_by(Site,sample_type,Sampling_Day)%>%
  ggplot(aes(x=Sampling_Day,y=HOBO_lumens,fill=Site,shape = sample_type))+
  geom_point(size=3)+
  scale_shape_manual(values = c(21,22))+
  scale_fill_manual(values = site_colors, labels = site_lookup, breaks = names(site_lookup))+
  theme_bw()+
  facet_wrap(~Site)

temporal_long_wLimits_outliersRmv%>%
  group_by(Site,sample_type)%>%
  ggplot(aes(x=Site,y=HOBO_lumens,fill=Site,shape = sample_type))+
  geom_point(size = 3)+
  facet_wrap(~sample_type,scales = "free")+
  scale_fill_manual(values = site_colors, labels = site_lookup, breaks = names(site_lookup))+
  scale_shape_manual(values = c(21,22))+
  theme_bw()

temporal_long_wLimits_outliersRmv %>%
  group_by(sample_type, Site) %>%
  summarise(
    across(
      c(HOBO_lumens, HOBO_temp_F), 
      list(mean = ~mean(.x, na.rm = TRUE), sd = ~sd(.x, na.rm = TRUE)), 
      .names = "{.col}_{.fn}"
    )
  ) %>%
  mutate(
    HOBO_lumens = paste0(round(HOBO_lumens_mean, 2), " ± ", round(HOBO_lumens_sd, 2)),
    HOBO_temp_F = paste0(round(HOBO_temp_F_mean, 2), " ± ", round(HOBO_temp_F_sd, 2))
  ) %>%
  select(sample_type, Site, HOBO_lumens, HOBO_temp_F)

```

## NPOC and TN
```{r NPOC TN}

temporal_long_wLimits_outliersRmv%>%
  pivot_longer(cols = c(NPOC, TN), names_to = "water_chem", values_to = "concen_uM")%>%
  group_by(Site,sample_type)%>%
  ggplot(aes(x=Site,y=concen_uM,color=Site,linetype = sample_type))+
  geom_boxplot()+
  facet_wrap(~water_chem, scales="free",ncol=1)+
  theme_bw()+
  scale_color_manual(values = site_colors, labels = site_lookup, breaks = names(site_lookup))

temporal_long_wLimits_outliersRmv %>%
  group_by(sample_type, Site) %>%
  summarise(
    across(
      c(NPOC, TN), 
      list(mean = ~mean(.x, na.rm = TRUE), sd = ~sd(.x, na.rm = TRUE)), 
      .names = "{.col}_{.fn}"
    )
  ) %>%
  mutate(
    NPOC = paste0(round(NPOC_mean, 2), " ± ", round(NPOC_sd, 2)),
    TN = paste0(round(TN_mean, 2), " ± ", round(TN_sd, 2))
  ) %>%
  select(sample_type, Site, NPOC, TN)


```
## Flow cytometry
```{r flow cytrometry}

temporal_long_wLimits_outliersRmv%>%
  pivot_longer(cols = c(Prochlorococcus, Synechococcus,Unpigmented_cells,Picoeukaryotes), names_to = "fcm_data", values_to = "counts")%>%
  group_by(Site,sample_type)%>%
  select(Site,sample_type,fcm_data,counts)%>%
  drop_na()%>%
  ggplot(aes(x=Site,y=counts,color=Site,linetype = sample_type))+
  geom_boxplot()+
  facet_wrap(~fcm_data, scales="free",ncol=1)+
  theme_bw()+
  scale_color_manual(values = site_colors, labels = site_lookup, breaks = names(site_lookup))

temporal_long_wLimits_outliersRmv %>%
  group_by(sample_type, Site) %>%
  summarise(
    across(
      c(Prochlorococcus, Synechococcus, Unpigmented_cells, Picoeukaryotes), 
      list(mean = ~mean(.x, na.rm = TRUE), sd = ~sd(.x, na.rm = TRUE)), 
      .names = "{.col}_{.fn}"
    )
  ) %>%
  mutate(
    Prochlorococcus = paste0(round(Prochlorococcus_mean, 2), " ± ", round(Prochlorococcus_sd, 2)),
    Synechococcus = paste0(round(Synechococcus_mean, 2), " ± ", round(Synechococcus_sd, 2)),
    Unpigmented_cells = paste0(round(Unpigmented_cells_mean, 2), " ± ", round(Unpigmented_cells_sd, 2)),
    Picoeukaryotes = paste0(round(Picoeukaryotes_mean, 2), " ± ", round(Picoeukaryotes_sd, 2))
  ) %>%
  select(sample_type, Site, Prochlorococcus, Synechococcus, Unpigmented_cells, Picoeukaryotes)

```

## Metabolite counts and concentration ranges
```{r mtab summary}

temporal_long_wLimits_outliersRmv %>%
  group_by(Site, sample_type, metabolites) %>%
  summarise(mean_concentration = mean(concentration), LOD = first(LOD), .groups = "drop") %>%
  group_by(Site, sample_type) %>%
  summarise(count_above_LOD = sum(mean_concentration > LOD), .groups = "drop")

```

#Statistical tests
## Bray-Curtis Dissimilarity Calculations
```{r bray curtis dissimilitaries}

dist_usvi_mtab.d <- vegan::vegdist(temporal.mat_outliersRmv_log2, method = "bray", binary = FALSE, na.rm = TRUE)

dist_usvi_mtab.df <- vegan::betadisper(dist_usvi_mtab.d, temporal_wide_outliersRmv_log2$Site) %>%
  purrr::pluck("distances") %>%
  tibble::enframe(value = "dissimilarity", name = "adaptedDervLabel") %>%
  dplyr::left_join(., (temporal_wide_outliersRmv_log2 %>%
                         dplyr::select(adaptedDervLabel, sample_type, Date, Time, Sampling_Day, Site, rep_group) %>%
                         droplevels),
                   by = c("adaptedDervLabel" = "adaptedDervLabel")) %>%
  droplevels

g1 <- print(
  ggplot(data = dist_usvi_mtab.df)
  + theme_bw()
  + geom_boxplot(aes(x = Site, y = dissimilarity, 
                     # color = Site, 
                     # color = sampling_time, 
                     group = interaction(Site, sample_type)), 
                 color = "black",
                 position = position_dodge2(padding = 0.2, preserve = "single"),
                 show.legend = FALSE, outliers = FALSE)
  + geom_point(aes(x = Site, y = dissimilarity, fill = Site, group = interaction(Site, sample_type), shape = sample_type), 
               alpha = 1.0, size = 3, position = position_jitterdodge(dodge.width = 0.75, seed = 48105, jitter.width = 0.2))
  + scale_shape_manual(values = c(22, 21, 23), labels = c(sampling_time_lookup, "NA"), breaks = c(names(sampling_time_lookup), NA))
  # + scale_color_manual(values = sampling_time_colors, labels = sampling_time_lookup, breaks = names(sampling_time_lookup))
  # + scale_color_manual(values = site_colors, labels = site_lookup, breaks = names(site_lookup))
  + scale_fill_manual(values = site_colors, labels = site_lookup, breaks = names(site_lookup))
  + scale_y_continuous(expand = expansion(mult = c(0,0.1)), name = "Bray-Curtis dissimilarity")
  + scale_x_discrete(labels = site_lookup, name = "Site")
  +labs(title = "A) Targeted Metabolome")
  + theme(axis.title = element_text(size = 12, face = "bold", colour = "grey30"),
          panel.background = element_blank(), panel.border = element_rect(fill = "NA", colour = "grey30"),
          panel.grid = element_blank(),
          legend.position = "right",
          legend.key = element_blank(),
          legend.title = element_text(size = 12, face = "bold", colour = "grey30"),
          legend.text = element_text(size = 12, colour = "grey30"))
  + guides(color = "none",
           fill = guide_legend(order = 2, ncol = 1, title = "Site", direction = "vertical",
                               override.aes = list(color = "black", stroke = 1, shape = 21, size = 2)),
           shape = guide_legend(order = 1, ncol = 1, title = "Sampling time", direction = "vertical",
                                override.aes = list(color = "black", stroke = 1, size = 2)))
)

#ggsave("/Volumes/KujLab/Brianna/Projects/2022_0328_CINAR_BC/BC_CoralSeagrass_Temporal/USVI_2021_CINAR_temporal_R_Script/figures/BrayCurtis_Dissimilarity_TargetedMetabolome_20241217.png",plot = last_plot(),height=4.25,width=6.5,units = "in",dpi=600)
```

### Bray-Curtis Dissimilarity Statistical Comparison within sites across time
```{r bray curtis dissimilarity stats}

# Function to perform Kruskal-Wallis test and FDR adjustment for a given site
KW_samplingTime_dissimiliarty <- function(data, site_filter) {
  data %>%
    ungroup() %>%
    filter(Site == site_filter) %>%
    summarise(
      Kruskal_p = kruskal.test(dissimilarity ~ sample_type)$p.value,
      .groups = "drop")
}

# List of subsets with corresponding labels
subset_list <- list(
  kw_SG = list(Site = "LB_seagrass"),
  kw_TK = list(Site = "Tektite"),
  kw_YZ = list(Site = "Yawzi")
)

# Combine all results into one summary table
set.seed(123)
summary_table <- purrr::imap_dfr(
  subset_list,
  ~ KW_samplingTime_dissimiliarty(dist_usvi_mtab.df, .x$Site) %>%
      mutate(Site = .x$Site) # Use .x$Site to extract the site name
)

# View the summary table
print(summary_table)

```
### Bray-Curtis Dissimilarity Statistical Comparison between sites within time
```{r KW and Wilcoxon test across sites within time}

# Function to perform Kruskal-Wallis test and FDR adjustment for a given sampling time
run_kruskal_time_fdr <- function(data, sample_type_filter) {
  data %>%
    ungroup() %>%
    filter(sample_type == sample_type_filter) %>%
    summarise(
      Kruskal_p = kruskal.test(dissimilarity ~ Site)$p.value,
      .groups = "drop"
    ) %>%
    mutate(
      FDR_adjusted_p = p.adjust(Kruskal_p, method = "fdr"),
      sig_05_fdr = FDR_adjusted_p < 0.05
    )
}

# Function to perform pairwise Wilcoxon rank-sum tests for a given sampling time
run_pairwise_tests <- function(data, sample_type_filter) {
  # Filter data for the specific sampling time
  filtered_data <- data %>%
    filter(sample_type == sample_type_filter)
  
  # Ensure 'dissimilarity' and 'Site' are vectors
  pairwise_result <- pairwise.wilcox.test(
    x = filtered_data$dissimilarity,
    g = filtered_data$Site,
    p.adjust.method = "fdr"
  )
  
  # Convert results to a tidy dataframe
  broom::tidy(pairwise_result) %>%
    mutate(samplingTime = sample_type_filter)
}

# List of subsets with corresponding labels
subset_list <- list(
  kw_dawn = list(samplingTime = "dawn"),
  kw_afternoon = list(samplingTime = "peak_photo")
)

# Run Kruskal-Wallis tests
summary_table <- purrr::imap_dfr(
  subset_list,
  ~ run_kruskal_time_fdr(dist_usvi_mtab.df, .x$samplingTime) %>%
      mutate(samplingTime = .x$samplingTime) 
)

# Run pairwise Wilcoxon tests for significant Kruskal-Wallis results
pairwise_results <- purrr::imap_dfr(
  subset_list,
  ~ run_pairwise_tests(dist_usvi_mtab.df, .x$samplingTime)
)

# View the results
print(summary_table)  # Kruskal-Wallis test results
print(pairwise_results)  # Pairwise Wilcoxon test results

```

## PERMANOVA
### Categorical variables - does sampling day matter? 
```{r PERMANOVA sampling day}

permanova_dm <- temporal_wide_outliersRmv_log2%>%
  select(adaptedDervLabel, 
         Site, 
         Sampling_Day,
         sample_type,
         c("2'deoxycytidine":"valine")
         )%>%
  drop_na() # Remove rows with NaN values before the transformation

permanova.mat <- permanova_dm%>%
  select(c("2'deoxycytidine":"valine"))%>%
  as.matrix()

rownames(permanova.mat) <- permanova_dm$adaptedDervLabel

set.seed(123)

result_combined <- adonis2(
  permanova.mat ~ Site + sample_type + Sampling_Day,
  data = permanova_dm,
  method = "bray",
  permutations = 999,
  by = "terms"
)

print(result_combined)

# Extract results from the combined PERMANOVA model
results_df <- data.frame(
  Factor = rownames(result_combined),  # Automatically uses the factor names from the combined model
  F_value = result_combined$F,        # Extract F-values
  R2 = result_combined$R2,            # Extract R² values
  P_value = result_combined$`Pr(>F)`  # Extract p-values
)

# Format the results_df for manuscript presentation
results_df %>%
  mutate(
    F_value = round(F_value, 2),
    R2 = round(R2, 3),
    P_value = ifelse(P_value < 0.001, "<0.001", round(P_value, 3))
  ) %>%
  kable(
    format = "html",  # Change to "latex" for LaTeX documents
    col.names = c("Factor", "F Value", "R²", "P Value"),
    caption = "Table 1. Targeted metabolome PERMANOVA results for environmental and biological variables",
    align = "l"  # Center align all columns
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "center"
  ) %>%
  column_spec(2:4, bold = TRUE) %>%
  add_header_above(c(" " = 1, "PERMANOVA Statistics" = 3))

```

## CAPSCALE model of continuous environmental & biological variables
```{r linear model}

#Log2 transform continuous variables to make them on the same general scale
cleanData <- temporal_wide_outliersRmv_log2 %>%
  select(adaptedDervLabel, Site, sample_type, NPOC, TN, Prochlorococcus, Synechococcus, Picoeukaryotes,Unpigmented_cells, c("2'deoxycytidine":valine)) %>%
  mutate(across(c(NPOC:Unpigmented_cells), ~log2(. + 1))) %>%
  drop_na() #removal of rows that have NA

#Check for co-linearity of variables
cor(cleanData[, c("Prochlorococcus", "Synechococcus", "Picoeukaryotes", "NPOC", "TN")])

cleanData.mat <- cleanData%>%
  select(c("2'deoxycytidine":"valine"))%>%
  as.matrix()

rownames(cleanData.mat) <- cleanData$adaptedDervLabel

#Calculate bray curtis dsitnace of metabolome data 
set.seed(123)
bray_dist <- vegan::vegdist(cleanData.mat, method = "bray", binary = FALSE, na.rm = TRUE)

#Run RDA using capscale
set.seed(123)
db_rda <- capscale(bray_dist ~ Site + sample_type + Prochlorococcus + Picoeukaryotes + NPOC + TN, data = cleanData)
summary(db_rda)

#Calculate significance with an ANOVA
set.seed(123)
anova(db_rda, by = "terms")

```
## Kruskal-Wallis
### Kruskal-Wallis: Within Site/Sampling Time but across days
Since metabolite samples were measured at the same three sites, at the same time of day (dawn and afternoon) every day over four days, we need to test if the metabolite concentrations within a site/sampling time (e.g., Tektite - dawn) vary over those four days. If they do, metabolites will need to be further investigate on a day-by-day basis. If not, then metabolites can be grouped by their sampling time across the four days, thus increasing our statistical power when investigating differences between sampling time. 
```{r KW across days}

# Function to perform Kruskal-Wallis test and FDR adjustment for a given subset of data
run_kruskal_fdr <- function(data, site_filter, sample_type_filter) {
  data %>%
    ungroup() %>%
    filter(Site == site_filter, sample_type == sample_type_filter) %>%
    group_by(metabolites) %>%
    summarise(
      Kruskal_p = kruskal.test(log2_concen ~ Sampling_Day)$p.value,
      .groups = "drop"
    ) %>%
    mutate(
      FDR_adjusted_p = p.adjust(Kruskal_p, method = "fdr"),
      sig_05_fdr = FDR_adjusted_p < 0.05
    )
}

# List of subsets with corresponding labels
subset_list <- list(
  kw_SG_peak = list(site = "LB_seagrass", sample_type = "peak_photo"),
  kw_SG_dawn = list(site = "LB_seagrass", sample_type = "dawn"),
  kw_TK_peak = list(site = "Tektite", sample_type = "peak_photo"),
  kw_TK_dawn = list(site = "Tektite", sample_type = "dawn"),
  kw_YZ_peak = list(site = "Yawzi", sample_type = "peak_photo"),
  kw_YZ_dawn = list(site = "Yawzi", sample_type = "dawn")
)

# Combine all results into one summary table
set.seed(123)
summary_table <- purrr::imap_dfr(
  subset_list,
  ~ run_kruskal_fdr(temporal_long_wLimits_outliersRmv, .x$site, .x$sample_type) %>%
      mutate(site_sample = .y)
)

# View the summary table
print(summary_table)

```

#### Count number of significant metabolites
```{r count_significant_metabolites, echo=FALSE}

# Calculate the count of significant metabolites
num_significant <- summary_table %>%
  filter(FDR_adjusted_p < 0.05) %>%
  nrow()

# Output the count
cat("The number of metabolites with an adjusted p-value (FDR) < 0.05 within a Site/Sampling Time but across days is:", num_significant)

```

### Kruskal-Wallis: Within Site between sampling time (dawn and afternoon)
Since metabolite concentrations did not vary significantly post-FDR correction, metabolites can be grouped by their site and sampling time and tested for signficant differences in concentration between dawn and afternoon. 

```{r KW sampling time}

# Function to perform Kruskal-Wallis test and FDR adjustment for a given site
KW_samplingTime_fdr <- function(data, site_filter) {
  data %>%
    ungroup() %>%
    filter(Site == site_filter) %>%
    group_by(metabolites) %>%
    summarise(
      Kruskal_p = kruskal.test(log2_concen ~ sample_type)$p.value,
      .groups = "drop"
    ) %>%
    mutate(
      FDR_adjusted_p = p.adjust(Kruskal_p, method = "fdr"),
      sig_05_fdr = FDR_adjusted_p < 0.05
    )
}

# List of subsets with corresponding labels
subset_list <- list(
  kw_SG = list(site = "LB_seagrass"),
  kw_TK = list(site = "Tektite"),
  kw_YZ = list(site = "Yawzi")
)

# Combine all results into one summary table
set.seed(123)
summary_table <- purrr::imap_dfr(
  subset_list,
  ~ KW_samplingTime_fdr(temporal_long_wLimits_outliersRmv, .x$site) %>%
      mutate(site = .x$site) # Use .x$site to extract the site name
)

# View the summary table
print(summary_table)

sig_FDR_mtabs <- summary_table%>%
  filter(FDR_adjusted_p < 0.05)

```

#### Boxplots - significant metabolites dawn vs. afternoon
```{r plot significant mtabs}

x <- sig_FDR_mtabs%>%
  select(metabolites)%>%
  deframe() 

temporal_long_wLimits_outliersRmv%>%
  filter(metabolites %in% x)%>%
  ggplot(aes(x=Site,y=log2_concen,color=Site,linetype = sample_type),linewidth =2)+
  geom_boxplot()+
  facet_wrap(~metabolites,scales= "free")+
  scale_shape_manual(values = c(21,22))+
  scale_color_manual(values = CINAR_temporal_palette)+
  theme_bw(base_size = 12)


```

#### Heatpmap - Significant metaboltites dawn vs. afternoon
```{r sig mtab heatmaps}

# Create individual heatmaps for each group

x <- sig_FDR_mtabs%>%
  filter(site == "Tektite")%>%
  select(metabolites)%>%
  deframe()

heatmap1 <- temporal_long_wLimits_outliersRmv%>%
  filter(Site == "Tektite", metabolites %in% x )%>%
  mutate(date_time = factor(date_time))%>%
  group_by(metabolites,date_time,sample_type)%>%
  mutate(median_conc = round(median(concentration),3),
         forHeatmap = if_else(median_conc < LOD, NA, median_conc))%>%
  ggplot(aes(date_time, metabolites, fill = log2(forHeatmap+1))) +
  geom_tile(color="black", aes(fill = forHeatmap)) +
  scale_fill_viridis(trans = 'log', na.value = "white", direction = 1,name = NULL,
                     labels = label_number(accuracy = 0.001))+ # Round to 2 decimal places  
  theme_minimal(base_size = 14)+
  theme(legend.position = "right", 
        text = element_text(colour = "black"),
        axis.title.x = element_blank(), 
        axis.text.x = element_blank(), 
        axis.ticks.x = element_blank(),
        axis.title.y = element_blank(),
        plot.margin = margin(10, 10, 10, 10),
        panel.border = element_rect(fill = NA, color = "black", size = 1),
        panel.grid = element_blank())         # Remove grid lines

x <- sig_FDR_mtabs%>%
  filter(site == "Yawzi")%>%
  select(metabolites)%>%
  deframe()

heatmap2 <- temporal_long_wLimits_outliersRmv%>%
  filter(Site == "Yawzi", metabolites %in% x )%>%
  mutate(date_time = factor(date_time))%>%
  group_by(metabolites,date_time,sample_type)%>%
  mutate(median_conc = round(median(concentration),3),
         forHeatmap = if_else(median_conc < LOD, NA, median_conc))%>%
  ggplot(aes(date_time, metabolites, fill = log2(forHeatmap+1))) +
  geom_tile(color="black", aes(fill = forHeatmap)) +
  scale_fill_viridis(trans = 'log', na.value = "white", direction = 1,name = NULL,
                     labels = label_number(accuracy = 0.001))+ # Round to 2 decimal places  
  theme_minimal(base_size = 14)+
    theme(legend.position = "right", 
        text = element_text(colour = "black"),
        axis.title.x = element_blank(), 
        axis.text.x = element_blank(), 
        axis.ticks.x = element_blank(),
        axis.title.y = element_blank(),
        plot.margin = margin(10, 10, 10, 10),
        panel.border = element_rect(fill = NA, color = "black", size = 1),
        panel.grid = element_blank())         # Remove grid lines

x <- sig_FDR_mtabs%>%
  filter(site == "LB_seagrass")%>%
  select(metabolites)%>%
  deframe()

heatmap3 <- temporal_long_wLimits_outliersRmv%>%
  filter(Site == "LB_seagrass", metabolites %in% x )%>%
  mutate(date_time = factor(date_time))%>%
  group_by(metabolites,date_time,sample_type)%>%
  mutate(median_conc = round(median(concentration),3),
         forHeatmap = if_else(median_conc < LOD, NA, median_conc))%>%
  ggplot(aes(date_time, metabolites, fill = log2(forHeatmap+1))) +
  geom_tile(color="black", aes(fill = forHeatmap)) +
  scale_fill_viridis(trans = 'log', na.value = "white", direction = 1,name = NULL,
                     labels = label_number(accuracy = 0.001))+ # Round to 2 decimal places  
  theme_minimal(base_size = 14)+
  labs(x="Sampling Day")+
      theme(legend.position = "right", 
        text = element_text(colour = "black"),
        axis.title.y = element_blank(),
        plot.margin = margin(10, 10, 10, 10),
        panel.border = element_rect(fill = NA, color = "black", size = 1),
        panel.grid = element_blank())+         # Remove grid lines
  scale_x_discrete(labels=c('Day1','Day1',
                            'Day2','Day2',
                            'Day3','Day3',
                            'Day4','Day4'))

# Create the small dawn/afternoon heatmap
dawn_heatmap <- temporal_long_wLimits_outliersRmv%>%
    mutate(date_time = factor(date_time))%>%
  ggplot(aes(x = date_time, y = 1, fill = sample_type)) +
  geom_tile() +
  scale_fill_manual(values = c("dawn" = "black", "peak_photo" = "white")) +
  theme_bw() + # Remove all grid, text, and axes
    theme(legend.position = "none", 
        text = element_text(colour = "black"),
        axis.title.x = element_blank(), 
        axis.text.x = element_blank(), 
        axis.ticks.x = element_blank(),
        axis.title.y = element_blank(), 
        axis.text.y = element_blank(), 
        axis.ticks.y = element_blank(),
        plot.margin = margin(2, 2, 2, 2),
        panel.border = element_rect(fill = NA, color = "black", size = 1),
        panel.grid = element_blank())         # Remove grid lines    

# Combine the heatmaps with white space in between
combined_plot <- heatmap1 / heatmap2 / heatmap3 / dawn_heatmap+ 
  plot_layout(ncol = 1, heights = c(1, 1, 1,0.075)) & 
  plot_annotation(theme = theme(plot.background = element_rect(fill = "white", color = NA)))

# Show the final plot
combined_plot

#ggsave("/Volumes/KujLab/Brianna/Projects/2022_0328_CINAR_BC/BC_CoralSeagrass_Temporal/USVI_2021_CINAR_temporal_R_Script/figures/Temporally_Significant_TargetedMetabolites_log10_concentration_20241217.png",plot = last_plot(),height=8.5,width=11,units = "in",dpi=600)

```

# Manuscript figures
## Figure 1. Map of St. John USVI
```{r figure 1}
# 
# # Raw shapefiles
# usa <- st_read("./Analysis/R/scripts/USVI_map/stanford-vt021tk4894-shapefile/","vt021tk4894")
# sttstj <- st_read("./Analysis/R/scripts/USVI_map/stsj_fin")
# nps <- st_read("./Analysis/R/scripts/USVI_map/NPS_-_Land_Resources_Division_Boundary_and_Tract_Data_Service")
# 
# # site metadata
# 
# temporal_sites <- read.table("./BC_CoralSeagrass_Temporal/temporal_sites.txt", sep = "\t", header = TRUE)
# temporal_sites$Site <- factor(temporal_sites$Site, levels = c( "Yawzi", "Tektite","LB_seagrass"))
# usvi <- usa %>% filter(state == "United States Virgin Islands")
# 
# reefZONE <- sttstj %>%
#   filter(ZONE %in% c("Forereef", "Reef Crest", "Backreef"))
# 
# vinp <- nps %>% filter(PARKNAME == "Virgin Islands")
# 
# #plot
# temporal_map <- ggplot() +
#   geom_sf(data = usvi, fill = "#e5dccd",color = "#6699CC") +
#   geom_sf(data = reefZONE, fill = "#cc6677", color = "#882255",alpha=0.5) +
#   geom_sf(data = vinp, fill = NA, color = "#117733", linewidth = 0.6) +
#   coord_sf(xlim = c(-64.8040, -64.655), ylim = c(18.2895, 18.378364), expand = FALSE) +
#   geom_point(data = temporal_sites, mapping = aes(x = Lon, y = Lat, fill = Site),colour="#333333", pch = 21, size = 3,alpha=0.9) +
#   scale_fill_manual(values = CINAR_temporal_palette) +
#   annotation_scale(location = "bl", width_hint = 0.2) +
#   theme(panel.grid.major = element_blank(), panel.background = element_rect(fill = "#6699CC"),legend.position = c(0.902, 0.21))+
#   labs(x = "Longitude",y="Latitude")+
#   geom_text() +
#   annotate("text", label = "St. John,\nU.S. Virgin Islands",x = -64.74, y = 18.34, size = 4, colour = "black",fontface=2)+
#   annotate("text", label = "Coral reef habitat",x = -64.785, y = 18.307, size = 3, colour = "#882255",fontface=2)+
#   annotate("text", label = "Virgin Islands \nNational Park",x = -64.715, y = 18.3585, size = 3, colour = "#117733",fontface=2)+
#   annotate("text", label = "Lameshur Bay",x = -64.735, y = 18.305, size = 3, colour = "black",fontface=2)
# 
# temporal_map + geom_segment(aes(x = -64.735, y = 18.307, xend = -64.728, yend = 18.312),
#                             arrow = arrow(length = unit(0.01, "npc")))

```

```{r figure 1 zoom}
# 
# #plot
# temporal_map_zoom <- ggplot() +
#   geom_sf(data = usvi, fill = "#e5dccd",color = "#6699CC") +
#   geom_sf(data = reefZONE, fill = "#cc6677", color = "#882255",alpha=0.5) +
#   geom_sf(data = vinp, fill = NA, color = "#117733", linewidth = 0.6) +
#   coord_sf(xlim = c(-64.727, -64.72), ylim = c(18.308, 18.320), expand = FALSE) +
#   geom_point(data = temporal_sites, mapping = aes(x = Lon, y = Lat, fill = Site),colour="#333333", pch = 21, size = 4,alpha=0.88) +
#   scale_fill_manual(values = CINAR_temporal_palette) +
#   annotation_scale(location = "bl", width_hint = 0.2) +
#   theme(panel.grid.major = element_blank(), panel.background = element_rect(fill = "#6699CC"),legend.position = "top",
#   axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
#   labs(x = "Longitude",y="Latitude")+
#   geom_text() +
#   annotate("text", label = "Great Lameshur Bay",x = -64.723, y = 18.315, size = 2.5, colour = "black",fontface=2)
# 
# temporal_map_zoom

```

```{r save figure 1}

#save
#png(file = "./BC_CoralSeagrass_Temporal/figures/CINAR_temporal_USVI_StJohn_Sampling_Map.png",width = 8.5, height = 7, units = "in", res = 300)

#save
#png(file = "./BC_CoralSeagrass_Temporal/figures/CINAR_temporal_USVI_StJohn_Sampling_Map_zoom.png",width = 6, height = 5.5, units = "in", res = 300)

```

```{r removal map variables}
# 
# rm(vinp,usvi,usa,sttstj,temporal_sites,reefZONE,nps)

```

## Figure 2.
```{r figure 2}

```
## Figure 3. 
```{r figure 3}

```
## Figure 4.
```{r figure 4}

```